---
title: "Technical Replicates Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(here)
library(dplyr)
library(stringr)
library(phyloseq)
library(vegan)
library(Polychrome)
library(RColorBrewer)
library(ggplot2)
```

## Loading the data

Load the metadata file:

```{r load-metadata}
#| cache: true
#| label: load-metadata
metadata <- read.delim(here::here("data", "metadata_bacteria.tsv"))
```

Now we will group the technical replicates together. We will use the `sample_group` column to group the samples. The `sample_group` column is a unique identifier for the technical replicates. We will remove the last letter from the `sampleID` column to get a new name for the sample.

```{r group-technical-replicates}
#| cache: true
#| label: group-technical-replicates
# get a new name for sample 
metadata <- metadata %>%
  dplyr::mutate(sample_name = stringr::str_replace(sampleID, "[a-zA-Z]$", "")) %>%
  dplyr::group_by(sample_group) %>%
  dplyr::mutate(sample_name = first(sample_name)) %>%
  dplyr::ungroup() %>%
  dplyr::select(sampleID, sample_name, everything()) %>%
  as.data.frame()

# set rownames
rownames(metadata) <- metadata$sampleID

# check the new metadata
metadata
```

Now read the phyloseq object:

```{r load-phyloseq}
#| cache: true
#| label: load-phyloseq
phyloseq_object <- readRDS(here::here("results-bacteria/phyloseq/dada2_phyloseq.rds"))

# replace the metadata
sample_data(phyloseq_object) <- metadata

# execute some useful grouping on metadata
sample_data(phyloseq_object)$group <- as.factor(sample_data(phyloseq_object)$sample_group)

# Create a new grouping variable using only the last 20 characters of 'condition' and 'date'
sample_data(phyloseq_object)$date_condition <- as.factor(with(
  sample_data(phyloseq_object), 
  paste0(
    gsub(" ", "", substr(condition, nchar(condition) - 12, nchar(condition))), 
    "_", 
    gsub(" ", "", date)
  )
))
```

## Checking samples

Try to see sequencing depths and other information about the samples:

```{r sequencing-depth}
#| cache: true
#| label: sequencing-depth
# calculate sequencing depth for each samples
sequencing_depth <- phyloseq::sample_sums(phyloseq_object)

# get data from samples
samples_data <- data.frame(sample_data(phyloseq_object))

# add sequencing depth to samples data
samples_data$sequencing_depth <- sequencing_depth

# Create the bar plot
depth_plot <- ggplot2::ggplot(
  samples_data, 
  aes(x = sampleID, y = sequencing_depth, fill = sample_name)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

### Calculate the distance matrix

We will calculate the distance matrix using the `bray-curtis method:

```{r distance-matrix}
#| cache: true
#| label: distance-matrix
# calculate a distance matrix
distance_matrix <- phyloseq::distance(phyloseq_object, method = "bray")
```


### Permanova

We will use the `adonis` function from the `vegan` package to test if the technical replicates are similar to each other. We will use the `sample_group` column as the strata.

```{r permanova}
#| cache: true
#| label: permanova
# now doing the permanova
adonis_result <- vegan::adonis2(distance_matrix ~ group, data = as(sample_data(phyloseq_object), "data.frame"))
print(adonis_result)
```

The **PERMANOVA** (Permutational Multivariate Analysis of Variance) output provides insights into the differences in community composition across the groups in the dataset. Here’s a detailed interpretation of each section of the output:

#### Degrees of Freedom (Df)

-   **Model (12)**: This indicates that there are 12 levels in the grouping variable `group` (13 groups - 1 = 12).
-   **Residual (26)**: The remaining 26 degrees of freedom represent the unexplained variance within groups.
-   **Total (38)**: The total degrees of freedom in the dataset.

#### Sum of Squares (SumOfSqs)

-   **Model (14.3595)**: This is the sum of squares for the grouping factor, representing the variability explained by `group`.
-   **Residual (0.3871)**: This is the residual sum of squares, indicating the variability within groups.
-   **Total (14.7466)**: The total sum of squares for the dataset.

#### R-squared (R2)

-   **Model (0.97375)**: This value shows that 97.4% of the total variability is explained by differences between the groups, indicating a very high proportion of variation accounted for by the grouping variable.
-   **Residual (0.02625)**: Only 2.6% of the variability is left unexplained by the model, which suggests that samples within the same group are quite similar.

#### F-statistic (F) and P-value (Pr(\>F))

-   **F (80.378)**: The F-statistic is high, indicating a strong separation between groups relative to within-group variability.
-   **Pr(\>F) (0.001)**: The p-value of 0.001 is highly significant (denoted by `***`), suggesting a statistically significant difference in community composition across the groups.

#### Summary

The PERMANOVA test reveals that:

-   The `group` variable explains a significant portion of the variation in community composition (97.4%).
-   The high F-value and low p-value (0.001) indicate that this difference is statistically significant.
-   The result supports the hypothesis that groups have distinct community compositions, with minimal overlap in terms of their ecological or compositional profiles.

### PCoA plot

**Principal Coordinates Analysis (PCoA)** is an ordination technique used to visualize patterns in complex, high-dimensional data by projecting samples into a lower-dimensional space based on a distance matrix.

#### Key Points

- **Input**: PCoA takes a **distance or dissimilarity matrix** as input, such as Bray-Curtis, Jaccard, or Euclidean distances.
- **Goal**: It represents the relative distances between samples in a reduced number of dimensions, typically 2D or 3D, for easy visualization.
- **Output**: Samples that are closer together in the PCoA plot are more similar, while those farther apart are more dissimilar.
- **Interpretation**: The axes (principal coordinates) explain portions of the variance in the data, helping reveal clusters, trends, or group separations.

#### Applications

- **Microbiome and Ecology**: Useful for visualizing community composition differences.
- **Genomics and Transcriptomics**: Reveals patterns in gene expression across conditions.

#### PCoA vs. PCA

- **PCoA** uses a distance matrix, making it ideal for data with non-linear relationships or compositional constraints.
- **PCA** works directly with raw data and assumes linear relationships.

#### PCoA plot by technical replicates

Now we will plot a PCoA plot to visualize the technical replicates.

```{r pcoa}
#| cache: true
#| label: pcoa
# Calculate the PCoA
pca_ordination <- phyloseq::ordinate(phyloseq_object, method = "PCoA", distance = "bray")
```

In the following pictures we are grouping by technical replicates

```{r pcoa-plot}
#| cache: true
#| label: pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(sample_data(phyloseq_object)$group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
pcoa_plot <- phyloseq::plot_ordination(phyloseq_object, pca_ordination, color = "group") +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on distance metrics (Bray-Curtis)") +
  ggplot2::labs(color = "group") +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Technical Replicates")) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(pcoa_plot)
```

#### PCoA plot by date and condition (sample replicates)

Visualize the same plot but grouping with date and condition (which should group sample replicates)

```{r}
#| cache: true
#| label: pcoa-date-condition
# Count the number of unique levels in the new grouping variable
num_levels <- length(unique(sample_data(phyloseq_object)$date_condition))

# Plot the PCoA
pcoa_plot <- phyloseq::plot_ordination(phyloseq_object, pca_ordination, color = "date_condition") +
  ggplot2::geom_point(size = 4, alpha = 0.1) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(color = NULL) +  # Remove the legend title
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(pcoa_plot)
```

### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

#### Key Features of NMDS

- **Non-metric Approach**: NMDS emphasizes the rank order of dissimilarities, allowing it to handle data that do not meet the assumptions of linearity or normality.

- **Dimensionality Reduction**: It projects high-dimensional data into a lower-dimensional space (typically 2D or 3D) to facilitate visualization and interpretation.

- **Stress Value**: NMDS iteratively adjusts the configuration of points to minimize a measure called "stress," which quantifies the disparity between the distances in the reduced-dimensional space and the original dissimilarities. Lower stress values indicate a better fit.

#### Interpreting an NMDS Plot

- **Proximity of Points**: Samples plotted closer together are more similar in composition, while those farther apart are more dissimilar.

- **Axes**: The axes in an NMDS plot are arbitrary and do not correspond to specific gradients or variables. The orientation and rotation of the plot can change without affecting the interpretation of the relationships between points.

- **Stress Value**: A stress value below 0.2 is generally considered acceptable, indicating that the plot provides a reasonable representation of the data.

#### Applications of NMDS

- **Ecology**: Visualizing species composition across different sites or conditions.

- **Microbial Community Analysis**: Comparing microbial communities based on sequencing data.

- **Environmental Science**: Assessing the impact of environmental gradients on community structure.

NMDS is a flexible and robust method for exploring complex datasets, especially when the data do not conform to the assumptions required by other ordination techniques. 

```{r nmds}
#| cache: true
#| label: nmds
#| output: false
# calculate the NMDS
nmds_ordination <- phyloseq::ordinate(phyloseq_object, method = "NMDS", distance = "bray")
```
#### NMDS plot by technical replicates

Let's plot the NMDS plot focusing on the technical replicates

```{r}
#| cache: true
#| label: nmds-plot
# count the number of levels in the group column
num_levels <- length(unique(sample_data(phyloseq_object)$group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot <- phyloseq::plot_ordination(phyloseq_object, nmds_ordination, color = "group") +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("NMDS Plot on distance metrics (Bray-Curtis)") +
  ggplot2::labs(color = "group") +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Technical Replicates")) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(nmds_plot)
```

#### NMDS plot by date and condition (sample replicates)

Now, let's visualize the NMDS plot but grouping with date and condition 
(which should group sample replicates)

```{r}
#| cache: true
#| label: nmds-date-condition
# Count the number of unique levels in the new grouping variable
num_levels <- length(unique(sample_data(phyloseq_object)$date_condition))

# Plot the NMDS
nmds_plot <- phyloseq::plot_ordination(phyloseq_object, nmds_ordination, color = "date_condition") +
  ggplot2::geom_point(size = 4, alpha = 0.1) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("NMDS Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(color = NULL) +  # Remove the legend title
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(nmds_plot)
```
