---
title: "Technical Replicates Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)

store_path <- here::here("_targets")
```

## Loading the data

Load data from bacteria samples. We have grouped the technical replicates 
together. We will use the `sample_group` column to group the samples. 
The `sample_group` column is a unique identifier for the technical replicates. 
We have removed the last letter from the `sampleID` column to get a new name 
for the sample:

```{r load-metadata}
#| label: load-metadata
# load data from targets
targets::tar_load(metadata, store = store_path)

# check the new metadata
metadata
```

Load phyloseq object other stuff

```{r load-phyloseq}
#| label: load-phyloseq
targets::tar_load(phyloseq_object, store = store_path)
targets::tar_load(otu_table_matrix, store = store_path)
targets::tar_load(samples_data, store = store_path)
```

## Checking samples

### Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
# Create the bar plot
depth_plot <- ggplot2::ggplot(
  samples_data, 
  aes(x = sampleID, y = sequencing_depth, fill = sample_name)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

### Attempt to estimate the sequencing depth

Try to follow the suggestion of Patrick D. Schloss to estimate the sequencing depth
(see [How to find the best sampling depth for rarefaction (CC202)](https://youtu.be/c7H8jLjTxSE?si=Wu5JLRCsBJuRyZc7) on
[Riffomonas Project](https://www.youtube.com/@Riffomonas)).

```{r estimate-sampling-coverage}
#| label: estimate-sampling-coverage
sampling_coverage <- samples_data %>%
  rownames_to_column(var = "rowname") %>%
  select(-everything(), sampleID, sequencing_depth) %>%
  arrange(sequencing_depth) %>%
  rename(n_seqs = sequencing_depth)
sampling_coverage
```

Plot the sampling coverage: here's the histogram of the number of sequences per sample:

```{r coverage-histogram}
#| label: coverage-histogram
sampling_coverage %>%
  ggplot(aes(x=n_seqs)) +
  geom_histogram(binwidth=10000) +
  coord_cartesian(xlim=c(0, 100000))
```

A jitter plot of the number of sequences per sample (plot random x values 
to avoid overplotting):

```{r coverage-jitter}
#| label: coverage-jitter
sampling_coverage %>%
  ggplot(aes(x=1, y=n_seqs)) +
  geom_jitter() +
  scale_y_log10()
```

A lineplot of the number of sequences per sample:

```{r}
#| label: coverage-lineplot
sampling_coverage %>%
  arrange(n_seqs) %>%
  ggplot(aes(x=1:nrow(.), y=n_seqs))+
  geom_line()
```

We didn't see a curve like the one we see in the tutorial. If we choose a 
sequencing depth of 6,000, we will throw away only 3 samples.

Calculate the *good's coverage* as defined by the formula:

$$
C_{Good's} = 100(1-\frac{n_1}{N_T}) 
$$

where $n_1$ is the number of OTUs that are observed only once 
and $N_T$ is the total number of sequences.

```{r goods-coverage}
#| label: goods-coverage
targets::tar_load(coverage_stats, store = store_path)

coverage_stats %>% 
  arrange(goods) %>%
  print(n=10)
```

Here's the plot of the Good's coverage:

```{r goods-coverage-plot}
#| label: goods-coverage-plot
coverage_stats %>%
  ggplot(aes(x=n_seqs, y=goods)) +
  geom_point()
```

The three samples with the lowest sampling depth have a Good's coverage above 
99.6%. We don't see the need of throwing them away.

### Calculate the distance matrix

Load the distance matrix using the *bray-curtis* method using a
*rarefaction* method as suggested by Patrick D. Schloss:

```{r distance-matrix}
#| label: distance-matrix
targets::tar_load(distance_matrix, store = store_path)
```

### Permanova

We will use the `adonis2` function from the `vegan` package to test if the technical
replicates are similar to each other. The formula should include the `sample_group` 
as a factor to account for biological variability and `technical_rep` to assess 
technical replicate similarity:

```{r permanova}
#| label: permanova
# load the permanova with targets
targets::tar_load(adonis_result, store = store_path)
print(adonis_result)
```

The **PERMANOVA** (Permutational Multivariate Analysis of Variance) output provides
insights into the differences in community composition across the groups in the
dataset. Here’s a detailed interpretation of each section of the output:

#### Degrees of Freedom (Df)

- **Model**: This indicates that there are 12 levels in the grouping 
  variable `sample_group` (13 groups - 1 = 12) and 2 for technical replicates.
- **Residual (24)**: The remaining 24 degrees of freedom represent the
  unexplained variance within groups.
- **Total (38)**: The total degrees of freedom in the dataset.

#### Sum of Squares (SumOfSqs)

- **Model**: This is the sum of squares for the grouping factor, representing 
  the variability explained by both variables.
- **Residual**: This is the residual sum of squares, indicating the variability
  within groups.
- **Total**: The total sum of squares for the dataset.

#### R-squared (R2)

- **Model**: This value shows that the reported percentage of the total 
  variability is explained by differences between the groups, indicating a very 
  high proportion of variation accounted for `sample_group`. Vice versa, the
  `technical_rep` variable explains a very low proportion of the variation.
- **Residual**: Only this percentage of the variability is left unexplained by 
  the model (i.e., within groups).

#### F-statistic (F) and P-value (Pr(\>F))

- **F**: The F-statistic is high for `sample_group`, indicating a strong separation 
  between groups relative to within-group variability. `technical_rep` has a low F-value,
  suggesting that technical replicates are similar to each other.
- **Pr(\>F)**: The p-value of 0.001 is highly significant (denoted by `***`) for
  `sample_group`, indicating that the observed differences are unlikely to have
  occurred by chance. The p-value for `technical_rep` is not significant.

#### Summary

- `sample_group`: The highly significant p-value (0.001) and high R² (99.53%) 
  indicate that differences between biological samples (sample_name) account 
  for the majority of the variance in your data.
- `technical_rep`: The non-significant p-value (0.234) and very low R² (0.044%) 
  suggest that differences among technical replicates (technical_rep) do not 
  contribute significantly to the variance. This implies that technical 
  replicates are consistent with each other, as expected.

### PCoA plot

**Principal Coordinates Analysis (PCoA)** is an ordination technique used to visualize patterns in complex, high-dimensional data by projecting samples into a lower-dimensional space based on a distance matrix.

#### Key Points

- **Input**: PCoA takes a **distance or dissimilarity matrix** as input, such as Bray-Curtis, Jaccard, or Euclidean distances.
- **Goal**: It represents the relative distances between samples in a reduced number of dimensions, typically 2D or 3D, for easy visualization.
- **Output**: Samples that are closer together in the PCoA plot are more similar, while those farther apart are more dissimilar.
- **Interpretation**: The axes (principal coordinates) explain portions of the variance in the data, helping reveal clusters, trends, or group separations.

#### Applications

- **Microbiome and Ecology**: Useful for visualizing community composition differences.
- **Genomics and Transcriptomics**: Reveals patterns in gene expression across conditions.

#### PCoA vs. PCA

- **PCoA** uses a distance matrix, making it ideal for data with non-linear 
  relationships or compositional constraints. Is a dimensional reduction technique
- **PCA** works directly with raw data and assumes linear relationships. Is not
  used in microbiome studies since it doesn't deal with 0 (see Pat Schloss YT channel)

#### PCoA plot by technical replicates

Now we will plot a PCoA plot to visualize the technical replicates.

```{r pcoa}
#| label: pcoa
# retrieve the PCoA object from targets
targets::tar_load(pcoa_object, store = store_path)

pcoa_results <- pcoa_object$pcoa_results
percent_explained <- pcoa_object$percent_explained
pcoa_df <- pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r pcoa-plot}
#| label: pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by date and condition (sample replicates)

Visualize the same plot but grouping with date and condition (which should group sample replicates)

```{r}
#| label: pcoa-date-condition
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = date_condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal 
components in a PCA or PCoA analysis. It helps visualize the proportion of variance 
explained by each principal component, allowing you to determine the optimal 
number of components to retain.

```{r}
tibble(pe = pcoa_results$values$Cum_corr_eig * 100,
       axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line()
```

### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

#### Key Features of NMDS

- **Non-metric Approach**: NMDS emphasizes the rank order of dissimilarities, allowing it to handle data that do not meet the assumptions of linearity or normality.

- **Dimensionality Reduction**: It projects high-dimensional data into a lower-dimensional space (typically 2D or 3D) to facilitate visualization and interpretation.

- **Stress Value**: NMDS iteratively adjusts the configuration of points to minimize a measure called "stress," which quantifies the disparity between the distances in the reduced-dimensional space and the original dissimilarities. Lower stress values indicate a better fit.

#### Interpreting an NMDS Plot

- **Proximity of Points**: Samples plotted closer together are more similar in composition, while those farther apart are more dissimilar.

- **Axes**: The axes in an NMDS plot are arbitrary and do not correspond to specific gradients or variables. The orientation and rotation of the plot can change without affecting the interpretation of the relationships between points.

- **Stress Value**: A stress value below 0.2 is generally considered acceptable, indicating that the plot provides a reasonable representation of the data.

#### Applications of NMDS

- **Ecology**: Visualizing species composition across different sites or conditions.

- **Microbial Community Analysis**: Comparing microbial communities based on sequencing data.

- **Environmental Science**: Assessing the impact of environmental gradients on community structure.

NMDS is a flexible and robust method for exploring complex datasets, especially when the data do not conform to the assumptions required by other ordination techniques. 

```{r nmds}
targets::tar_load(nmds_object)

nmds_tb <- nmds_object$nmds_tb
```
#### NMDS plot by technical replicates

Let's plot the NMDS plot focusing on the technical replicates

```{r}
#| label: nmds-plot
# count the number of levels in the group column
num_levels <- length(unique(metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x=NMDS1, y=NMDS2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

#### NMDS plot by date and condition (sample replicates)

Now, let's visualize the NMDS plot but grouping with date and condition 
(which should group sample replicates)

```{r}
#| label: nmds-date-condition

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x=NMDS1, y=NMDS2, color = date_condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```
