---
title: "Technical Replicates Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
editor: visual
---

```{r setup}
#| label: setup
#| output: false
library(here)
library(dplyr)
library(stringr)
library(phyloseq)
library(vegan)
library(Polychrome)
library(ggplot2)
```

## Loading the data

Load the metadata file:

```{r load-metadata}
#| cache: true
#| label: load-metadata
metadata <- read.delim(here::here("data", "metadata_bacteria.tsv"))
```

Now we will group the technical replicates together. We will use the `sample_group`
column to group the samples. The `sample_group` column is a unique identifier for
the technical replicates. We will remove the last letter from the `sampleID` column
to get a new name for the sample.

```{r group-technical-replicates}
#| cache: true
#| label: group-technical-replicates
# get a new name for sample 
metadata <- metadata %>%
  dplyr::mutate(sample_name = stringr::str_replace(sampleID, "[a-zA-Z]$", "")) %>%
  dplyr::group_by(sample_group) %>%
  dplyr::mutate(sample_name = first(sample_name)) %>%
  dplyr::ungroup() %>%
  dplyr::select(sampleID, sample_name, everything())
metadata
```

Now read the phyloseq object:

```{r load-phyloseq}
#| cache: true
#| label: load-phyloseq
phyloseq_object <- readRDS(here::here("results-bacteria/phyloseq/dada2_phyloseq.rds"))
```

## Testing the technical replicates

### Permanova

We will use the `adonis` function from the `vegan` package to test if the technical
replicates are similar to each other. We will use the `sample_group` column as the
strata.

```{r permanova}
#| cache: true
#| label: permanova
# calculate a distance matrix
distance_matrix <- phyloseq::distance(phyloseq_object, method = "bray")

# now doing the permanova
sample_data(phyloseq_object)$group <- as.factor(sample_data(phyloseq_object)$sample_group)
adonis_result <- vegan::adonis2(distance_matrix ~ group, data = as(sample_data(phyloseq_object), "data.frame"))
print(adonis_result)
```

The **PERMANOVA** (Permutational Multivariate Analysis of Variance) output provides insights into the differences in community composition across the groups in the dataset. Here’s a detailed interpretation of each section of the output:

#### Degrees of Freedom (Df)
  - **Model (12)**: This indicates that there are 12 levels in the grouping variable `group`
  (13 groups - 1 = 12).
  - **Residual (26)**: The remaining 26 degrees of freedom represent the unexplained variance within groups.
  - **Total (38)**: The total degrees of freedom in the dataset.

#### Sum of Squares (SumOfSqs)
  - **Model (14.3595)**: This is the sum of squares for the grouping factor, representing the variability explained by `group`.
  - **Residual (0.3871)**: This is the residual sum of squares, indicating the variability within groups.
  - **Total (14.7466)**: The total sum of squares for the dataset.

#### R-squared (R2)
  - **Model (0.97375)**: This value shows that 97.4% of the total variability is explained by differences between the groups, indicating a very high proportion of variation accounted for by the grouping variable.
  - **Residual (0.02625)**: Only 2.6% of the variability is left unexplained by the model, which suggests that samples within the same group are quite similar.

#### F-statistic (F) and P-value (Pr(>F))
  - **F (80.378)**: The F-statistic is high, indicating a strong separation between groups relative to within-group variability.
  - **Pr(>F) (0.001)**: The p-value of 0.001 is highly significant (denoted by `***`), suggesting a statistically significant difference in community composition across the groups.

#### Summary
The PERMANOVA test reveals that:

  - The `group` variable explains a significant portion of the variation in community composition (97.4%).
  - The high F-value and low p-value (0.001) indicate that this difference is statistically significant.
  - The result supports the hypothesis that groups have distinct community compositions, with minimal overlap in terms of their ecological or compositional profiles.

### PCoA plot

Now we will plot a PCoA plot to visualize the technical replicates.

```{r pcoa}
#| cache: true
#| label: pcoa
# Calculate the PCoA
ordination <- ordinate(phyloseq_object, method = "PCoA", distance = "bray")

# count the number of levels in the group column
num_levels <- length(unique(sample_data(phyloseq_object)$group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
pcoa_plot <- phyloseq::plot_ordination(phyloseq_object, ordination, color = "group") +
  ggplot2::geom_point(size = 4) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on distance metrics (Bray-Curtis)") +
  ggplot2::labs(color = "group") +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Sample Group"))

print(pcoa_plot)
```

```{r}
# Create a new grouping variable with newline characters in long labels
# Create a new grouping variable using only the last 20 characters of 'condition' and 'date'
sample_data(phyloseq_object)$group_combined <- with(
  sample_data(phyloseq_object), 
  paste0(substr(condition, nchar(condition) - 19, nchar(condition)), "_", date)
)

# Count the number of unique levels in the new grouping variable
num_levels <- length(unique(sample_data(phyloseq_object)$group_combined))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Calculate the PCoA using the new grouping variable for color
ordination <- ordinate(phyloseq_object, method = "PCoA", distance = "bray")

# Plot the PCoA
pcoa_plot <- phyloseq::plot_ordination(phyloseq_object, ordination, color = "group_combined") +
  ggplot2::geom_point(size = 4) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(color = "Group (Condition_Date)") +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Sample Group")) +
  ggplot2::theme(legend.position = "bottom")  # Move the legend to the bottom

print(pcoa_plot)
```