---
title: "Reactor vs Algae"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
    embed-resources: true  # Embed all resources directly into the HTML file
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(ampvis2)

store_path <- here::here("stores", "reactor_vs_algae")
```

## Load and select data

Load data and select only the stuff I need:

```{r load_data}
#| label: load_data
# load objects from targets
targets::tar_load(phyloseq_subset, store = store_path)
targets::tar_load(metadata_subset, store = store_path)
```

## Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
#| description: Display the sequencing depth of each sample.
# load stuff from store directory
tar_load(samples_data, store = store_path)

# Create the bar plot
depth_plot <- ggplot2::ggplot(
  samples_data,
  aes(x = sampleID, y = sequencing_depth, fill = sample_name)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

## Barplots

Let's create some barplots to visualize the relative abundance of taxa in the
dataset. Here are the available taxonomic ranks in the dataset:

```{r taxonomic-ranks}
#| label: taxonomic-ranks
#| description: List the available taxonomic ranks in the dataset.
phyloseq::rank_names(phyloseq_subset)
```

### By Phylum

Create a barplot of the relative abundance Phylum in the dataset. Sort samples
by sampleID.

```{r barplot-phylum}
#| label: barplot-phylum
#| description: Create a barplot of the relative abundance of Phylum in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum"
#| fig-align: center
# melt by phylum
targets::tar_load(melted_phylum, store = store_path)

# make plot
taxrank <- "Phylum"
ggplot2::ggplot(melted_phylum, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = metadata_subset$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class

Create a barplot of the relative abundance Class in the dataset. Sort samples
by sampleID.

```{r barplot-class}
#| label: barplot-class
#| description: Create a barplot of the relative abundance of class in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class"
#| fig-align: center
# melt by class
targets::tar_load(melted_class, store = store_path)

# make plot
taxrank <- "Class"
ggplot2::ggplot(melted_class, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = metadata_subset$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmaps

### By phylum

Here are the heatmaps of the relative abundance of the top 10 most abundant
taxa at the Phylum level:

```{r heatmap-phylum}
#| label: heatmap-phylum
#| description: Create a heatmap of the relative abundance of the top 10 most abundant taxa at the Phylum level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 10 most abundant taxa at the Phylum level"
#| fig-align: center
targets::tar_load(heatmap_phylum, store = store_path)
print(heatmap_phylum)
```

### By class

Here are the heatmaps of the relative abundance of the top 20 most abundant
taxa at the Class level:

```{r heatmap-class}
#| label: heatmap-class
#| description: Create a heatmap of the relative abundance of the top 20 most abundant taxa at the Class level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 20 most abundant taxa at the Class level"
#| fig-align: center
targets::tar_load(heatmap_class, store = store_path)
print(heatmap_class)
```

## Samples Rarefaction

Rarefaction is a technique used to estimate species richness or diversity
in a community by standardizing the number of samples or reads across different
samples. It involves subsampling the data to a common depth, allowing for a fair
comparison of diversity metrics between samples with varying sequencing depths.

```{r rarefaction}
#| label: rarefaction
#| description: Perform rarefaction analysis on the OTU table
#|  and visualize the results.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Rarefaction curves with vegan"
# deal with rarefaction data
targets::tar_load(rarecurve_df, store = store_path)
targets::tar_load(rarefaction_depth, store = store_path)

# Add metadata to the rarecurve_df
rarecurve_df <- metadata_subset %>%
  select(sampleID, sample_name) %>%
  inner_join(rarecurve_df, by = c("sampleID" = "Site"))

# draw the plot
ggplot(rarecurve_df, aes(x = Sample, y = Species, color = sample_name, group = sampleID)) +
  geom_line() +
  geom_vline(xintercept = rarefaction_depth, linetype = "dashed", color = "red") +  # Add vertical line
  labs(
    title = "Rarefaction Curves for All Samples",
    x = "Number of Individuals Sampled",
    y = "Number of Species Observed"
  ) +
  theme_bw()
```

## Alpha diversity

Alpha diversity measures the diversity within a single sample or community.
It provides insights into the richness and evenness of species in a sample,
helping to compare diversity between samples or track changes over time.

### By sample_name

```{r alpha-diversity_sample_name}
#| label: alpha-diversity_sample_name
#| description: Calculate and visualize alpha diversity metrics for the samples.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Alpha diversity metrics for the samples"
tar_load(alpha_diversity_by_sample_name, store = store_path)
print(alpha_diversity_by_sample_name)
```

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on shannon metric:

```{r kruscal_shannon_sample_name}
#| label: kruscal_shannon_sample_name
#| description: Perform Kruskal-Wallis tests on the Shannon ~ sample_name
tar_load(kruscal_shannon_sample_name, store = store_path)
print(kruscal_shannon_sample_name)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test:

```{r dunn_shannon_sample_name}
#| label: dunn_shannon_sample_name
tar_load(dunn_shannon_sample_name, store = store_path)
dunn_shannon_sample_name$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

### By date_condition

```{r alpha-diversity_date_condition}
#| label: alpha-diversity_date_condition
#| description: Calculate and visualize alpha diversity metrics for date-condition.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Alpha diversity metrics for date-condition"
tar_load(alpha_diversity_by_date_condition, store = store_path)
print(alpha_diversity_by_date_condition)
```

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on shannon metric:

```{r kruscal_shannon_date_condition}
#| label: kruscal_shannon_date_condition
#| description: Perform Kruskal-Wallis tests on the Shannon ~ date_condition
tar_load(kruscal_shannon_date_condition, store = store_path)
print(kruscal_shannon_date_condition)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test:

```{r dunn_shannon_date_condition}
#| label: dunn_shannon_date_condition
tar_load(dunn_shannon_date_condition, store = store_path)
dunn_shannon_date_condition$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

## Beta diversity

### PCoA plot

#### PCoA plot by sample names

Now we will plot a PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r pcoa}
#| label: pcoa
# retrieve the PCoA object from targets
targets::tar_load(pcoa_object, store = store_path)

pcoa_results <- pcoa_object$pcoa_results
percent_explained <- pcoa_object$percent_explained
pcoa_df <- pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r pcoa-plot}
#| label: pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(metadata_subset$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by date and condition (sample replicates)

Visualize the same plot but grouping with date and condition (which should group sample replicates)

```{r pcoa-date-condition}
#| label: pcoa-date-condition
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = date_condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation between samples in reactor (pink and brown
on the left) and the other samples. The default *may24* and *nov23* cluster more or
less together (yellow and orange). *box_may24* (red) and *box_nov23* (blue) don't
cluster together and they are more simular to *box+dw_may24* (green) and *box+dw_nov23*
(purple).

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r scree-plot}
#| label: scree-plot
tibble(pe = pcoa_results$values$Cum_corr_eig * 100,
       axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  # add a red vertical line at y=2
  geom_vline(xintercept = 2, color = "red")
```

### NMDS plot

```{r nmds}
#| label: nmds
targets::tar_load(nmds_object, store = store_path)

nmds_tb <- nmds_object$nmds_tb
```
#### NMDS plot by sample names

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

```{r nmds-plot}
#| label: nmds-plot
# count the number of levels in the group column
num_levels <- length(unique(metadata_subset$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x=NMDS1, y=NMDS2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

#### NMDS plot by date and condition (sample replicates)

Now, let's visualize the NMDS plot but grouping with date and condition
(which should group sample replicates)

```{r nmds-date-condition}
#| label: nmds-date-condition

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x=NMDS1, y=NMDS2, color = date_condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```

### Between group distances

#### sample_name distances

Plot distances by sample names:

```{r sample_name_distances}
#| label: sample_name_distances
#| title: Distance matrix by sample names
#| description: Plot the distance matrix by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(bray_distance_by_sample_name_plot, store = store_path)
print(bray_distance_by_sample_name_plot)
```

#### date_condition distances

Plot distances by date_condition:

```{r date_condition_distances}
#| label: date_condition_distances
#| title: Distance matrix by date_condition
#| description: Plot the distance matrix by date_condition
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(bray_distance_by_date_condition_plot, store = store_path)
print(bray_distance_by_date_condition_plot)
```

### Permanova

#### sample_name permanova

Calculate beta dispersion for *sample_name* column:

```{r sample_name_beta_dispersion}
#| label: sample_name_beta_dispersion
#| title: Beta dispersion by sample names
#| description: Calculate beta dispersion by sample names
tar_load(sample_name_beta_dispersion, store = store_path)
print(sample_name_beta_dispersion$anova_result)
```

Since we don't see any significant differences in dispersion, we can proceed
with the PERMANOVA analysis. Plot beta dispersion:

```{r sample_name_beta_dispersion_plot}
#| label: sample_name_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(sample_name_beta_dispersion$beta_disp_df, aes(x = Group, y = Distance, fill = Group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Do permanova with *bray-curtis* on *sample_name* column:

```{r sample_name_permanova}
#| label: sample_name_permanova
# load the permanova with targets
targets::tar_load(sample_name_permanova, store = store_path)
print(sample_name_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r pairwise_sample_name_permanova}
#| label: pairwise_sample_name_permanova
# load the permanova with targets
targets::tar_load(pairwise_sample_name_permanova, store = store_path)

pairwise_sample_name_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```

No comparison has been filtered for *sample_name* column

#### date_condition permanova

Calculate beta dispersion for *date_condition* column:

```{r date_condition_beta_dispersion}
#| label: date_condition_beta_dispersion
#| title: Beta dispersion by sample names
#| description: Calculate beta dispersion by sample names
tar_load(date_condition_beta_dispersion, store = store_path)
print(date_condition_beta_dispersion$anova_result)
```

I see significant differences in dispersions: should I trust the `date_condition`
permanova? Plot beta dispersion:

```{r date_condition_beta_dispersion_plot}
#| label: date_condition_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(date_condition_beta_dispersion$beta_disp_df, aes(x = Group, y = Distance, fill = Group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```


Do permanova with *bray-curtis* on *date_condition* column:

```{r date_condition_permanova}
#| label: date_condition_permanova
# load the permanova with targets
targets::tar_load(date_condition_permanova, store = store_path)
print(date_condition_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r pairwise_date_condition_permanova}
#| label: pairwise_date_condition_permanova
# load the permanova with targets
targets::tar_load(pairwise_date_condition_permanova, store = store_path)

pairwise_date_condition_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```
