---
title: "Duckweed and box"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
    embed-resources: true  # Embed all resources directly into the HTML file
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(ampvis2)

source(here::here("R", "load_stuff.R"))
source(here::here("R", "misc.R"))
source(here::here("R", "alpha.R"))
source(here::here("R", "plots.R"))

store_path <- here::here("_targets")
```

## Load and select data

Load data and select only the stuff I need:

```{r load_data}
#| label: load_data
# load objects from targets
targets::tar_load(phyloseq_object, store = store_path)
targets::tar_load(metadata, store = store_path)

# select samples
samples_to_keep <- metadata %>% 
  dplyr::filter(date_condition %in% c(
    "default_nov23", "box_nov23", "default_may24", "box_may24", "box+dw_may24")) %>%
  select(sampleID)

# Subset the phyloseq object to include only the specified samples
phyloseq_subset <- phyloseq::prune_samples(sample_names(phyloseq_object) %in% t(samples_to_keep), phyloseq_object)

# Identify taxa with non-zero total abundance
non_zero_taxa <- taxa_sums(phyloseq_subset) > 0

# Prune taxa with zero abundance
phyloseq_pruned <- prune_taxa(non_zero_taxa, phyloseq_subset)

# Get otu table and rarefaction depth
otu_table_matrix <- get_otu_table(phyloseq_pruned)
rarefaction_depth <- min(sample_sums(phyloseq_pruned))

# get samples data
samples_data <- get_samples_data(phyloseq_pruned)

# get rid of unused objects
rm(phyloseq_object, phyloseq_subset, metadata, non_zero_taxa, samples_to_keep)

# get new metadata
metadata <- phyloseq::sample_data(phyloseq_pruned) %>%
  as_tibble() %>%
  dplyr::arrange(sample_number)
```

## Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
# Create the bar plot
depth_plot <- ggplot2::ggplot(
  samples_data,
  aes(x = sampleID, y = sequencing_depth, fill = sample_name)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

## Barplots

Let's create some barplots to visualize the relative abundance of taxa in the
dataset. Here are the available taxonomic ranks in the dataset:

```{r taxonomic-ranks}
#| label: taxonomic-ranks
#| description: List the available taxonomic ranks in the dataset.
phyloseq::rank_names(phyloseq_pruned)
```

### By Phylum

Create a barplot of the relative abundance Phylum in the dataset. Sort samples
by sampleID.

```{r barplot-phylum}
#| label: barplot-phylum
#| description: Create a barplot of the relative abundance of Phylum in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum"
#| fig-align: center
# melt by phylum
melted_phylum <- agglomerate_by_taxa(
  phyloseq_pruned,
  taxrank = "Phylum",
  sample_order = metadata$sampleID
)

# make plot
taxrank <- "Phylum"
ggplot2::ggplot(melted_phylum, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class

Create a barplot of the relative abundance Class in the dataset. Sort samples
by sampleID.

```{r barplot-class}
#| label: barplot-class
#| description: Create a barplot of the relative abundance of class in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class"
#| fig-align: center
# melt by class
melted_class <- agglomerate_by_taxa(
  phyloseq_pruned,
  taxrank = "Class",
  sample_order = metadata$sampleID
)

# make plot
taxrank <- "Class"
ggplot2::ggplot(melted_class, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmaps

```{r}
ampvis2_object <- phyloseq_to_ampvis2(phyloseq_pruned)
```

### By phylum

Here are the heatmaps of the relative abundance of the top 10 most abundant
taxa at the Phylum level:

```{r heatmap-phylum}
#| label: heatmap-phylum
#| description: Create a heatmap of the relative abundance of the top 10 most abundant taxa at the Phylum level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 10 most abundant taxa at the Phylum level"
#| fig-align: center
heatmap_phylum <- custom_heatmap(
  ampvis2_object,
  group_by = "date_condition",
  showRemainingTaxa = TRUE
)
print(heatmap_phylum)
```

### By class

Here are the heatmaps of the relative abundance of the top 20 most abundant
taxa at the Class level:

```{r heatmap-class}
#| label: heatmap-class
#| description: Create a heatmap of the relative abundance of the top 20 most abundant taxa at the Class level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 20 most abundant taxa at the Class level"
#| fig-align: center
heatmap_class <- custom_heatmap(
  ampvis2_object,
  group_by = "date_condition",
  showRemainingTaxa = TRUE,
  tax_add = "Class",
  tax_show = 20
)
print(heatmap_class)
```

## Samples Rarefaction

Rarefaction is a technique used to estimate species richness or diversity
in a community by standardizing the number of samples or reads across different
samples. It involves subsampling the data to a common depth, allowing for a fair
comparison of diversity metrics between samples with varying sequencing depths.

```{r rarefaction}
#| label: rarefaction
#| description: Perform rarefaction analysis on the OTU table
#|  and visualize the results.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Rarefaction curves with vegan"
# deal with rarefaction data
rarecurve_df <- calculate_rarecurve(otu_table_matrix)
rarefaction_depth <- min(sample_sums(phyloseq_pruned))

# Add metadata to the rarecurve_df
rarecurve_df <- metadata %>%
  select(sampleID, sample_name) %>%
  inner_join(rarecurve_df, by = c("sampleID" = "Site"))

# draw the plot
ggplot(rarecurve_df, aes(x = Sample, y = Species, color = sample_name, group = sampleID)) +
  geom_line() +
  geom_vline(xintercept = rarefaction_depth, linetype = "dashed", color = "red") +  # Add vertical line
  labs(
    title = "Rarefaction Curves for All Samples",
    x = "Number of Individuals Sampled",
    y = "Number of Species Observed"
  ) +
  theme_bw()
```
