---
title: "Bacteria - November 2023 Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
cache: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 5
    embed-resources: true  # Embed all resources directly into the HTML file
knitr:
  opts_chunk:
    dpi: 300
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(ampvis2)
library(openxlsx)

# where I can find artifacts
store_path <- here::here("stores", "november_2023-bacteria")

# a folder in which store data files
data_path <- here::here("analysis", "07-november_2023-bacteria")

if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  cat("Directory created:", data_path, "\n")
}

# set the default font size
theme_set(theme_minimal(base_size = 32))
theme_set(theme_bw(base_size = 32))
```

## Metadata

Let's load the metadata from the November 2023 experiments and try to do some analysis.

```{r load-metadata}
#| label: load-metadata
# load data from targets
targets::tar_load(nov_2023_metadata, store = store_path)

# write table for metadata
write.xlsx(nov_2023_metadata, file = here(data_path, "nov_2023-bacteria_metadata.xlsx"))

# check the new metadata
nov_2023_metadata
```

Download the [nov_2023-bacteria_metadata.xlsx](07-november_2023-bacteria/nov_2023-bacteria_metadata.xlsx) table.

```{r load-phyloseq}
#| label: load-phyloseq
# Load other phyloseq stuff
targets::tar_load(nov_2023_phyloseq_obj, store = store_path)
targets::tar_load(nov_2023_samples_data, store = store_path)

write.xlsx(nov_2023_samples_data, file = here(data_path, "nov_2023-bacteria_samples_data.xlsx"))
```

## Checking samples

### Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
# Create the bar plot
depth_plot <- ggplot2::ggplot(
  nov_2023_samples_data,
  aes(x = sampleID, y = sequencing_depth, fill = label)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  theme_minimal() +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads",
       fill = "Group") + # legend name for fill color
  theme(axis.text.x = element_text(
    angle = 90, 
    hjust = 1, 
    vjust = 0.5,
    margin = ggplot2::margin(b = 7)
  ))

# Display the plot
print(depth_plot)
```
Download the full [nov_2023-bacteria_samples_data.xlsx](07-november_2023-bacteria/nov_2023-bacteria_samples_data.xlsx) table.

### Permanova

Calculate the distance matrix using the *bray-curtis* method using a
*rarefaction* method as suggested by Patrick D. Schloss. Then
we will use the `adonis2` function from the `vegan` package to test if the technical
replicates are similar to each other. The formula should include the `sample_group`
as a factor to account for biological variability and `technical_rep` to assess
technical replicate similarity:

```{r tech_rep_permanova}
#| label: tech_rep_permanova
# load the permanova with targets
targets::tar_load(nov_2023_tech_rep_permanova, store = store_path)
print(nov_2023_tech_rep_permanova)
```

The bad score given by grouping samples by technical replicates should not explain
biological variability, discarding the hypothesis that biological replicates are
different.

## Barplots

Let's create some barplots to visualize the relative abundance of taxa in the
dataset. Here are the available taxonomic ranks in the dataset:

```{r taxonomic-ranks}
#| label: taxonomic-ranks
#| description: List the available taxonomic ranks in the dataset.
phyloseq::rank_names(nov_2023_phyloseq_obj)
```

### By Phylum

Create a barplot of the relative abundance Phylum in the dataset. Sort samples
by sampleID.

```{r barplot-phylum}
#| label: barplot-phylum
#| description: Create a barplot of the relative abundance of Phylum in the dataset.
#| fig-caption: "Barplot of Phylum"
#| fig-align: center
#| fig-width: 12
# load the melted data
tar_load(nov_2023_melted_phylum, store = store_path)
taxrank <- "Phylum"

# write table for melted_phylum
write.xlsx(nov_2023_melted_phylum, file = here(data_path, "nov_2023-bacteria_melted_phylum.xlsx"))

ggplot2::ggplot(nov_2023_melted_phylum, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$label) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
    angle = 90, 
    hjust = -1,
    vjust = 0.5, 
    margin = ggplot2::margin(b = 10)
  ))
```

Download the [nov_2023-bacteria_melted_phylum.xlsx](07-november_2023-bacteria/nov_2023-bacteria_melted_phylum.xlsx) table.

### By Phylum (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance
in the *Other taxa*

```{r barplot-phylum-grouped}
#| label: barplot-phylum-grouped
#| description: Create a barplot of the relative abundance of Phylum in the dataset grouping taxa with less than 5% abundance.
#| fig-caption: "Barplot of Phylum grouped (>5% abundance)"
#| fig-width: 12
taxrank <- "Phylum"
nov_2023_melted_phylum_grouped <- nov_2023_melted_phylum %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_phylum_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$label) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
    angle = 90,
    hjust = -1,
    vjust = 0.5,
    margin = ggplot2::margin(b = 10)
  ))
```

### By Class

Create a barplot of the relative abundance Class in the dataset. Sort samples
by sampleID.

```{r barplot-class}
#| label: barplot-class
#| description: Create a barplot of the relative abundance of class in the dataset.
#| fig-width: 15
#| fig-caption: "Barplot of Class"
#| fig-align: center
# load the melted data
tar_load(nov_2023_melted_class, store = store_path)
taxrank <- "Class"

# write table for melted_class
write.xlsx(nov_2023_melted_class, file = here(data_path, "nov_2023-bacteria_melted_class.xlsx"))

ggplot2::ggplot(nov_2023_melted_class, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$label) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
    angle = 90,
    hjust = -1,
    vjust = 0.5,
    margin = ggplot2::margin(b = 10)
  ))
```

Download the [nov_2023-bacteria_melted_class.xlsx](07-november_2023-bacteria/nov_2023-bacteria_melted_class.xlsx) table.

### By Class (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance

```{r barplot-class-grouped}
#| label: barplot-class-grouped
#| description: Create a barplot of the relative abundance of Class in the dataset grouping taxa with less than 5% abundance.
#| fig-caption: "Barplot of Class grouped (>5% abundance)"
#| fig-width: 12
taxrank <- "Class"
nov_2023_melted_class_grouped <- nov_2023_melted_class %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_class_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$label) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
    angle = 90,
    vjust = 0.5,
    hjust = -1,
    margin = ggplot2::margin(b = 10)
  ))
```

## Heatmaps

### By Phylum

Here are the heatmaps of the relative abundance of the top 10 most abundant
taxa at the Phylum level:

```{r heatmap-phylum}
#| label: heatmap-phylum
#| description: Create a heatmap of the relative abundance of the top 10 most abundant taxa at the Phylum level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 10 most abundant taxa at the Phylum level"
#| fig-align: center
tar_load(nov_2023_heatmap_phylum, store = store_path)

# write a file for heatmap_phylum
write.xlsx(nov_2023_heatmap_phylum$data, file = here(data_path, "nov_2023-bacteria_heatmap_phylum.xlsx"))

print(nov_2023_heatmap_phylum)
```

Download the [nov_2023-bacteria_heatmap_phylum.xlsx](07-november_2023-bacteria/nov_2023-bacteria_heatmap_phylum.xlsx) table.

### By Class

Here are the heatmaps of the relative abundance of the top 20 most abundant
taxa at the Class level:

```{r heatmap-class}
#| label: heatmap-class
#| description: Create a heatmap of the relative abundance of the top 20 most abundant taxa at the Class level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 20 most abundant taxa at the Class level"
#| fig-align: center
tar_load(nov_2023_heatmap_class, store = store_path)

# write a file for heatmap_class
write.xlsx(nov_2023_heatmap_class$data, file = here(data_path, "nov_2023-bacteria_heatmap_class.xlsx"))

print(nov_2023_heatmap_class)
```

Download the [nov_2023-bacteria_heatmap_class.xlsx](07-november_2023-bacteria/nov_2023-bacteria_heatmap_class.xlsx) table.

## Samples Rarefaction

Rarefaction is a technique used to estimate species richness or diversity
in a community by standardizing the number of samples or reads across different
samples. It involves subsampling the data to a common depth, allowing for a fair
comparison of diversity metrics between samples with varying sequencing depths.
The red line correspond to the lowest number of sequences present in samples: this
is the sequencing depth used in rarefaction step.

```{r rarefaction}
#| label: rarefaction
#| description: Perform rarefaction analysis on the OTU table
#|  and visualize the results.
#| fig-caption: "Rarefaction curves with vegan"
# load calculated rarefaction data
tar_load(nov_2023_rarecurve_df, store = store_path)
tar_load(nov_2023_rarefaction_depth, store = store_path)

# Add metadata to the rarecurve_df
nov_2023_rarecurve_df <- nov_2023_metadata %>%
  select(sampleID, sample_name, label) %>%
  inner_join(nov_2023_rarecurve_df, by = c("sampleID" = "Site"))

# draw the plot
ggplot(nov_2023_rarecurve_df, aes(x = Sample, y = Species, color = label, group = sampleID)) +
  geom_line() +
  geom_vline(xintercept = nov_2023_rarefaction_depth, linetype = "dashed", color = "red") +  # Add vertical line
  geom_text(
    aes(
      x = nov_2023_rarefaction_depth,
      y = max(Species) * 0.9,
      label = paste("Depth:", nov_2023_rarefaction_depth),
      angle = 90
    ),
    color = "red",
    vjust = -0.5
  ) +  # Add text annotation above the line
  labs(
    title = "Rarefaction Curves for All Samples",
    x = "Number of Individuals Sampled",
    y = "Number of Species Observed",
    color = "Group"
  ) +
  theme_bw()
```

## Alpha diversity

Alpha diversity measures the diversity within a single sample or community.
It provides insights into the richness and evenness of species in a sample,
helping to compare diversity between samples or track changes over time.

| Name       | Description                                                                                      |
|------------|--------------------------------------------------------------------------------------------------|
| **Observed**   | The total number of unique species observed in a community.                                   |
| **Shannon**    | Measures species diversity by taking into account both richness and evenness; higher values indicate more diversity. |
| **Simpson**    | Assesses the probability that two randomly selected individuals belong to the same species; lower values indicate higher diversity. |
| **InvSimpson** | The inverse of Simpson's index, providing a measure that increases with species richness and evenness; higher values indicate greater diversity. |
| **Fisher**     | A richness estimator that is particularly sensitive to rare species, providing a measure of species richness based on species abundance distribution. |
| **FaithPD**    | Measures phylogenetic diversity, calculating the total branch length of the phylogenetic tree representing the species present in a community. |

### By sample_name

```{r alpha-diversity_sample_name}
#| label: alpha-diversity_sample_name
#| description: Calculate and visualize alpha diversity metrics for the samples.
#| fig-caption: "Alpha diversity metrics for the samples"
tar_load(nov_2023_alpha_diversity_by_sample_name, store = store_path)
print(nov_2023_alpha_diversity_by_sample_name)
```

#### Shannon

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on *shannon* metric:

```{r kruscal_shannon_sample_name}
#| label: kruscal_shannon_sample_name
#| description: Perform Kruskal-Wallis tests on the Shannon ~ sample_name
tar_load(nov_2023_kruscal_shannon_sample_name, store = store_path)
print(nov_2023_kruscal_shannon_sample_name)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test to find significant
differences between pairs:

```{r dunn_shannon_sample_name}
#| label: dunn_shannon_sample_name
tar_load(nov_2023_dunn_shannon_sample_name, store = store_path)

# write the full table for dunn_shannon_sample_name
write.xlsx(nov_2023_dunn_shannon_sample_name$res, file = here(data_path, "nov_2023-bacteria_dunn_shannon_sample_name.xlsx"))

nov_2023_dunn_shannon_sample_name$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

Download the full [nov_2023-bacteria_dunn_shannon_sample_name.xlsx](07-november_2023-bacteria/nov_2023-bacteria_dunn_shannon_sample_name.xlsx) table.

### By label

```{r alpha-diversity_labels}
#| label: alpha-diversity_labels
#| description: Calculate and visualize alpha diversity metrics for labels.
#| fig-caption: "Alpha diversity metrics for label"
tar_load(nov_2023_alpha_diversity_by_labels, store = store_path)
print(nov_2023_alpha_diversity_by_labels)
```

#### Shannon

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on shannon metric:

```{r kruscal_shannon_labels}
#| label: kruscal_shannon_labels
#| description: Perform Kruskal-Wallis tests on the Shannon ~ label
tar_load(nov_2023_kruscal_shannon_labels, store = store_path)
print(nov_2023_kruscal_shannon_labels)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test to find significant
differences between pairs:

```{r dunn_shannon_labels}
#| label: dunn_shannon_labels
tar_load(nov_2023_dunn_shannon_labels, store = store_path)

# write the full table for dunn_shannon_labels
write.xlsx(nov_2023_dunn_shannon_labels$res, file = here(data_path, "nov_2023-bacteria_dunn_shannon_labels.xlsx"))

nov_2023_dunn_shannon_labels$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

Download the full [nov_2023-bacteria_dunn_shannon_labels.xlsx](07-november_2023-bacteria/nov_2023-bacteria_dunn_shannon_labels.xlsx) table.

## Beta Diversity Metrics

Bray-Curtis, Jaccard, and UniFrac metrics are commonly used measures in ecological
and microbiological studies to compare the similarity or dissimilarity between
communities based on species composition or abundance. Here’s a brief description
and summary of each:

| **Metric**          | **Description** |
|---------------------|-----------------|
| [**Bray-Curtis**](#bray-curtis) | Measures the compositional dissimilarity between samples based on abundance counts. |
| [**Jaccard**](#jaccard)| Measures the proportion of shared species (OTUs/ASVs) between two samples. Ignores abundance. |
| [**Unweighted UniFrac**](#unifrac) | Considers shared branch length between communities in a phylogenetic tree but ignores abundance. |
| [**Weighted UniFrac**](#wunifrac) | Similar to unweighted UniFrac but also accounts for species' relative abundances, giving more weight to dominant taxa. |

Let's represent each metric in a section

### Bray-Curtis {#bray-curtis}

#### PCoa Plots

**Principal Coordinates Analysis (PCoA)** is an ordination technique used to visualize patterns in complex, high-dimensional data by projecting samples into a lower-dimensional space based on a distance matrix.

##### PCoA plot by sample names

Now we will plot a Bray-Curtis PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r bray-pcoa}
#| label: bray-pcoa
# retrieve the PCoA object from targets
targets::tar_load(nov_2023_bray_pcoa_object, store = store_path)

pcoa_results <- nov_2023_bray_pcoa_object$pcoa_results
percent_explained <- nov_2023_bray_pcoa_object$percent_explained
pcoa_df <- nov_2023_bray_pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r bray-pcoa-plot}
#| label: bray-pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

##### PCoA plot by labels (sample replicates)

Visualize the same plot but grouping by labels (which should group sample replicates)

```{r bray-pcoa-labels}
#| label: bray-pcoa-labels
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation from *reactor samples* from *default*.
*Box* samples are closer togheter but distant from *reactor* conditions.

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r bray-scree-plot}
#| label: bray-scree-plot
# Determine which cumulative eigenvalues to use
if ("Cum_corr_eig" %in% colnames(pcoa_results$values)) {
  cumulative_eig <- pcoa_results$values$Cum_corr_eig * 100
} else {
  cumulative_eig <- pcoa_results$values$Cumul_eig * 100
}

# Create a tibble for plotting
plot_data <- tibble(
  pe = cumulative_eig,
  axis = seq_along(cumulative_eig)
)

# Generate the plot
ggplot(plot_data, aes(x = axis, y = pe)) +
  geom_line() +
  geom_vline(xintercept = 2, color = "red") +
  labs(
    x = "Principal Coordinate Axis",
    y = "Cumulative Percentage of Variance Explained",
    title = "Scree Plot of PCoA (Bray-Curtis)",
  ) +
  theme_minimal()
```

#### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

```{r bray-nmds}
#| label: bray-nmds
targets::tar_load(nov_2023_bray_nmds_object, store = store_path)

# get the tibble object
nmds_tb <- nov_2023_bray_nmds_object$nmds_tb
jitter <- 0.05
```
##### NMDS plot by sample names

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

::: {.callout-warning}
I've received a warning regarding the stress
value, which is close to 0 and this can be determined by having not enough data,
even if this value means a good fit for the NMDS.
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

```{r bray-nmds-plot}
#| label: bray-nmds-plot
#| fig.width: 10
#| fig.height: 8

# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = sample_name, shape = condition)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL,
    shape = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

::: {.callout-tip}
This type of graph is more informative with indexes which use phylogenetic information:
take a look at [Unweighted UniFrac - NMDS plot by sample names](#unifrac-nmds-sample-names)
and [Weighted UniFrac - NMDS plot by sample names](#wunifrac-nmds-sample-names)
:::

##### NMDS plot by labels (sample replicates)

Now, let's visualize the NMDS plot but grouping by labels
(which should group sample replicates)

::: {.callout-warning}
I've received a warning regarding the stress
value, which is close to 0 and this can be determined by having not enough data,
even if this value means a good fit for the NMDS.
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

```{r bray-nmds-labels}
#| label: bray-nmds-labels

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = label)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```

::: {.callout-tip}
This type of graph is more informative with indexes which use phylogenetic information:
take a look at [Unweighted UniFrac - NMDS plot by labels](#unifrac-nmds-labels)
and [Weighted UniFrac - NMDS plot by labels](#wunifrac-nmds-labels)
:::

#### Between group distances

##### sample_name distances

Plot distances by sample names:

```{r bray-sample_name_distances}
#| label: bray-sample_name_distances
#| title: Distance matrix by sample names
#| description: Plot the distance matrix by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_bray_distance_by_sample_name, store = store_path)
write.xlsx(nov_2023_bray_distance_by_sample_name, file = here(data_path, "nov_2023-bacteria_bray_distance_by_sample_name.xlsx"))

tar_load(nov_2023_bray_distance_by_sample_name_plot, store = store_path)
print(nov_2023_bray_distance_by_sample_name_plot)
```

Download the full [nov_2023-bacteria_bray_distance_by_sample_name.xlsx](07-november_2023-bacteria/nov_2023-bacteria_bray_distance_by_sample_name.xlsx) table.

##### labels distances

Plot distances by labels:

```{r bray-labels_distances}
#| label: bray-labels_distances
#| title: Distance matrix by labels
#| description: Plot the distance matrix by labels
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_bray_distance_by_labels, store = store_path)
write.xlsx(nov_2023_bray_distance_by_labels, file = here(data_path, "nov_2023-bacteria_bray_distance_by_labels.xlsx"))

tar_load(nov_2023_bray_distance_by_labels_plot, store = store_path)
print(nov_2023_bray_distance_by_labels_plot)
```

Download the full [nov_2023-bacteria_bray_distance_by_labels.xlsx](07-november_2023-bacteria/nov_2023-bacteria_bray_distance_by_labels.xlsx) table.

#### Permanova

Here we will do some analyses on beta diversity. First, we evaluate beta dispersion
which is essential to ensure the validity of statistical tests, accurately
interpret differences in community composition, and select appropriate analytical
methods. Neglecting this step can lead to incorrect conclusions about the
ecological patterns present in your data. For unbalanced designs PERMANOVA and ANOSIM were

1. too liberal if the smaller group had greater dispersion, and
2. too conservative if the larger group had greater dispersion.

This result was especially so for ANOSIM (see [here](https://stats.stackexchange.com/a/316386/451801)
for more information)

##### sample_name permanova

Calculate beta dispersion for *sample_name* column:

```{r bray-sample_name_beta_dispersion}
#| label: bray-sample_name_beta_dispersion
#| title: Beta dispersion by sample names
#| description: Calculate beta dispersion by sample names
tar_load(nov_2023_bray_sample_name_beta_dispersion, store = store_path)
print(nov_2023_bray_sample_name_beta_dispersion$anova_result)
```

Since we don't see any significant differences in dispersion, we can proceed
with the PERMANOVA analysis. Plot beta dispersion:

```{r bray-sample_name_beta_dispersion_plot}
#| label: bray-sample_name_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(
    nov_2023_bray_sample_name_beta_dispersion$beta_disp_df,
    aes(x = Group, y = Distance, fill = Group)
  ) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "Bray-Curtis Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Do permanova with *bray-curtis* on *sample_name* column:

```{r bray-sample_name_permanova}
#| label: bray-sample_name_permanova
# load the permanova with targets
targets::tar_load(nov_2023_bray_sample_name_permanova, store = store_path)
print(nov_2023_bray_sample_name_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r bray-pairwise_sample_name_permanova}
#| label: bray-pairwise_sample_name_permanova
# load the permanova with targets
targets::tar_load(nov_2023_pairwise_bray_sample_name_permanova, store = store_path)
write.xlsx(nov_2023_pairwise_bray_sample_name_permanova, file = here(data_path, "nov_2023-bacteria_pairwise_bray_sample_name_permanova.xlsx"))

nov_2023_pairwise_bray_sample_name_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```

No comparison has been filtered for *sample_name* column.
Download the full [nov_2023-bacteria_pairwise_bray_sample_name_permanova.xlsx](07-november_2023-bacteria/nov_2023-bacteria_pairwise_bray_sample_name_permanova.xlsx) table.

##### labels permanova

Calculate beta dispersion for *labels* column:

```{r bray-labels_beta_dispersion}
#| label: bray-labels_beta_dispersion
#| title: Beta dispersion by labels
#| description: Calculate beta dispersion by labels
tar_load(nov_2023_bray_labels_beta_dispersion, store = store_path)
print(nov_2023_bray_labels_beta_dispersion$anova_result)
```

I see significant differences in dispersions: should I trust the `labels`
permanova? The point is also that groups are unbalanced (ie less elements in
*default* group than the others). Permanova (and also ANOSIM) may suffer in
this particular case. Plot beta dispersion:

```{r bray-labels_beta_dispersion_plot}
#| label: bray-labels_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(
    nov_2023_bray_labels_beta_dispersion$beta_disp_df,
    aes(x = Group, y = Distance, fill = Group)
  ) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "Bray-Curtis Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Do permanova with *bray-curtis* on *label* column:

```{r bray-labels_permanova}
#| label: bray-labels_permanova
# load the permanova with targets
targets::tar_load(nov_2023_bray_labels_permanova, store = store_path)
print(nov_2023_bray_labels_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r bray-pairwise_labels_permanova}
#| label: bray-pairwise_labels_permanova
# load the permanova with targets
targets::tar_load(nov_2023_pairwise_bray_labels_permanova, store = store_path)
write.xlsx(nov_2023_pairwise_bray_labels_permanova, file = here(data_path, "nov_2023-bacteria_pairwise_bray_labels_permanova.xlsx"))

nov_2023_pairwise_bray_labels_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```

Download the full [nov_2023-bacteria_pairwise_bray_labels_permanova.xlsx](07-november_2023-bacteria/nov_2023-bacteria_pairwise_bray_labels_permanova.xlsx) table.

### Jaccard {#jaccard}

#### PCoA plot by sample names

Now we will plot a Jaccard PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r jaccard-pcoa}
#| label: jaccard-pcoa
# retrieve the PCoA object from targets
targets::tar_load(nov_2023_jaccard_pcoa_object, store = store_path)

pcoa_results <- nov_2023_jaccard_pcoa_object$pcoa_results
percent_explained <- nov_2023_jaccard_pcoa_object$percent_explained
pcoa_df <- nov_2023_jaccard_pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r jaccard-pcoa-plot}
#| label: jaccard-pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Jaccard)") +
  ggplot2::labs(title = "PCoA of Jaccard Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by labels (sample replicates)

Visualize the same plot but grouping with labels (which should group sample replicates)

```{r jaccard-pcoa-labels}
#| label: jaccard-pcoa-labels
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Jaccard)") +
  ggplot2::labs(title = "PCoA of Jaccard Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation from *reactor samples* from *default*.
*Box* samples are closer togheter but distant from *reactor* conditions and closer
to *default* than reactor.

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r jaccard-scree-plot}
#| label: jaccard-scree-plot
# Determine which cumulative eigenvalues to use
if ("Cum_corr_eig" %in% colnames(pcoa_results$values)) {
  cumulative_eig <- pcoa_results$values$Cum_corr_eig * 100
} else {
  cumulative_eig <- pcoa_results$values$Cumul_eig * 100
}

# Create a tibble for plotting
plot_data <- tibble(
  pe = cumulative_eig,
  axis = seq_along(cumulative_eig)
)

# Generate the plot
ggplot(plot_data, aes(x = axis, y = pe)) +
  geom_line() +
  geom_vline(xintercept = 2, color = "red") +
  labs(
    x = "Principal Coordinate Axis",
    y = "Cumulative Percentage of Variance Explained",
    title = "Scree Plot of PCoA (Jaccard)",
  ) +
  theme_minimal()
```

#### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

```{r jaccard-nmds}
#| label: jaccard-nmds
targets::tar_load(nov_2023_jaccard_nmds_object, store = store_path)

# get the tibble object
nmds_tb <- nov_2023_jaccard_nmds_object$nmds_tb
jitter <- 0.05
```
##### NMDS plot by sample names

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

::: {.callout-warning}
I've received a warning regarding the stress
value, which is close to 0 and this can be determined by having not enough data,
even if this value means a good fit for the NMDS.
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

```{r jaccard-nmds-plot}
#| label: jaccard-nmds-plot
#| fig.width: 10
#| fig.height: 8

# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = sample_name, shape = condition)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Jaccard)",
    color = NULL,
    shape = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

::: {.callout-tip}
This type of graph is more informative with indexes which use phylogenetic information:
take a look at [Unweighted UniFrac - NMDS plot by sample names](#unifrac-nmds-sample-names)
and [Weighted UniFrac - NMDS plot by sample names](#wunifrac-nmds-sample-names)
:::

##### NMDS plot by labels (sample replicates)

Now, let's visualize the NMDS plot but grouping with labels
(which should group sample replicates)

::: {.callout-warning}
I've received a warning regarding the stress
value, which is close to 0 and this can be determined by having not enough data,
even if this value means a good fit for the NMDS.
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

```{r jaccard-nmds-labels}
#| label: jaccard-nmds-labels

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = label)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Jaccard)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```

::: {.callout-tip}
This type of graph is more informative with indexes which use phylogenetic information:
take a look at [Unweighted UniFrac - NMDS plot by labels](#unifrac-nmds-labels)
and [Weighted UniFrac - NMDS plot by labels](#wunifrac-nmds-labels)
:::

### Unweighted UniFrac {#unifrac}

#### PCoA plot by sample names

Now we will plot a Unweighted Unifrac PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r unifrac-pcoa}
#| label: unifrac-pcoa
# retrieve the PCoA object from targets
targets::tar_load(nov_2023_unifrac_pcoa_object, store = store_path)

pcoa_results <- nov_2023_unifrac_pcoa_object$pcoa_results
percent_explained <- nov_2023_unifrac_pcoa_object$percent_explained
pcoa_df <- nov_2023_unifrac_pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r unifrac-pcoa-plot}
#| label: unifrac-pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Unweighted Unifrac)") +
  ggplot2::labs(title = "PCoA of Unweighted Unifrac Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by labels (sample replicates)

Visualize the same plot but grouping with labels (which should group sample replicates)

```{r unifrac-pcoa-labels}
#| label: unifrac-pcoa-labels
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Weigthed Unifrac)") +
  ggplot2::labs(title = "PCoA of Unweighted Unifrac Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation from *reactor samples* from *default*.
*Box* samples are closer togheter but distant from *reactor* conditions and closer
to *default* than reactor.

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r unifrac-scree-plot}
#| label: unifrac-scree-plot
# Determine which cumulative eigenvalues to use
if ("Cum_corr_eig" %in% colnames(pcoa_results$values)) {
  cumulative_eig <- pcoa_results$values$Cum_corr_eig * 100
} else {
  cumulative_eig <- pcoa_results$values$Cumul_eig * 100
}

# Create a tibble for plotting
plot_data <- tibble(
  pe = cumulative_eig,
  axis = seq_along(cumulative_eig)
)

# Generate the plot
ggplot(plot_data, aes(x = axis, y = pe)) +
  geom_line() +
  geom_vline(xintercept = 2, color = "red") +
  labs(
    x = "Principal Coordinate Axis",
    y = "Cumulative Percentage of Variance Explained",
    title = "Scree Plot of PCoA (Unweighted Unifrac)",
  ) +
  theme_minimal()
```

#### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

```{r unifrac-nmds}
#| label: unifrac-nmds
targets::tar_load(nov_2023_unifrac_nmds_object, store = store_path)

# get the tibble object
nmds_tb <- nov_2023_unifrac_nmds_object$nmds_tb
```

##### NMDS plot by sample names {#unifrac-nmds-sample-names}

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

```{r unifrac-nmds-plot}
#| label: unifrac-nmds-plot
#| fig.width: 10
#| fig.height: 8

# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = sample_name, shape = condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Unweigthed Unifrac)",
    color = NULL,
    shape = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

##### NMDS plot by labels (sample replicates) {#unifrac-nmds-labels}

Now, let's visualize the NMDS plot but grouping with labels
(which should group sample replicates)

```{r unifrac-nmds-labels}
#| label: unifrac-nmds-labels

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Unweigthed Unifrac)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```

#### Between group distances

##### sample_name distances

Plot distances by sample names:

```{r unifrac-sample_name_distances}
#| label: unifrac-sample_name_distances
#| title: Distance matrix by sample names
#| description: Plot the distance matrix by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_unifrac_distance_by_sample_name, store = store_path)
write.xlsx(nov_2023_unifrac_distance_by_sample_name, file = here(data_path, "nov_2023-bacteria_unifrac_distance_by_sample_name.xlsx"))

tar_load(nov_2023_unifrac_distance_by_sample_name_plot, store = store_path)
print(nov_2023_unifrac_distance_by_sample_name_plot)
```

Download the full [nov_2023-bacteria_unifrac_distance_by_sample_name.xlsx](07-november_2023-bacteria/nov_2023-bacteria_unifrac_distance_by_sample_name.xlsx) table.

##### labels distances

Plot distances by labels:

```{r unifrac-labels_distances}
#| label: unifrac-labels_distances
#| title: Distance matrix by labels
#| description: Plot the distance matrix by labels
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_unifrac_distance_by_labels, store = store_path)
write.xlsx(nov_2023_unifrac_distance_by_labels, file = here(data_path, "nov_2023-bacteria_unifrac_distance_by_labels.xlsx"))

tar_load(nov_2023_unifrac_distance_by_labels_plot, store = store_path)
print(nov_2023_unifrac_distance_by_labels_plot)
```

Download the full [nov_2023-bacteria_unifrac_distance_by_labels.xlsx](07-november_2023-bacteria/nov_2023-bacteria_unifrac_distance_by_labels.xlsx) table.

#### Permanova

Here we will do some analyses on beta diversity. First, we evaluate beta dispersion
which is essential to ensure the validity of statistical tests, accurately
interpret differences in community composition, and select appropriate analytical
methods. Neglecting this step can lead to incorrect conclusions about the
ecological patterns present in your data. For unbalanced designs PERMANOVA and ANOSIM were

1. too liberal if the smaller group had greater dispersion, and
2. too conservative if the larger group had greater dispersion.

This result was especially so for ANOSIM (see [here](https://stats.stackexchange.com/a/316386/451801)
for more information)

##### sample_name permanova

Calculate beta dispersion for *sample_name* column:

```{r unifrac-sample_name_beta_dispersion}
#| label: unifrac-sample_name_beta_dispersion
#| title: Beta dispersion by sample names
#| description: Calculate beta dispersion by sample names
tar_load(nov_2023_unifrac_sample_name_beta_dispersion, store = store_path)
print(nov_2023_unifrac_sample_name_beta_dispersion$anova_result)
```

Since we don't see any significant differences in dispersion, we can proceed
with the PERMANOVA analysis. Plot beta dispersion:

```{r unifrac-sample_name_beta_dispersion_plot}
#| label: unifrac-sample_name_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(
    nov_2023_unifrac_sample_name_beta_dispersion$beta_disp_df,
    aes(x = Group, y = Distance, fill = Group)
  ) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "unifrac-Curtis Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Do permanova with *unifrac-curtis* on *sample_name* column:

```{r unifrac-sample_name_permanova}
#| label: unifrac-sample_name_permanova
# load the permanova with targets
targets::tar_load(nov_2023_unifrac_sample_name_permanova, store = store_path)
print(nov_2023_unifrac_sample_name_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r unifrac-pairwise_sample_name_permanova}
#| label: unifrac-pairwise_sample_name_permanova
# load the permanova with targets
targets::tar_load(nov_2023_pairwise_unifrac_sample_name_permanova, store = store_path)
write.xlsx(nov_2023_pairwise_unifrac_sample_name_permanova, file = here(data_path, "nov_2023-bacteria_pairwise_unifrac_sample_name_permanova.xlsx"))

nov_2023_pairwise_unifrac_sample_name_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```

No comparison has been filtered for *sample_name* column.
Download the full [nov_2023-bacteria_pairwise_unifrac_sample_name_permanova.xlsx](07-november_2023-bacteria/nov_2023-bacteria_pairwise_unifrac_sample_name_permanova.xlsx) table.

##### labels permanova

Calculate beta dispersion for *labels* column:

```{r unifrac-labels_beta_dispersion}
#| label: unifrac-labels_beta_dispersion
#| title: Beta dispersion by sample names
#| description: Calculate beta dispersion by sample names
tar_load(nov_2023_unifrac_labels_beta_dispersion, store = store_path)
print(nov_2023_unifrac_labels_beta_dispersion$anova_result)
```

I see significant differences in dispersions: should I trust the `labels`
permanova? The point is also that groups are unbalanced (ie less elements in
*default* group than the others). Permanova (and also ANOSIM) may suffer in
this particular case. Plot beta dispersion:

```{r unifrac-labels_beta_dispersion_plot}
#| label: unifrac-labels_beta_dispersion_plot
#| title: Beta dispersion plot by sample names
#| description: Plot beta dispersion by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
# Create the boxplot
ggplot(
    nov_2023_unifrac_labels_beta_dispersion$beta_disp_df,
    aes(x = Group, y = Distance, fill = Group)
  ) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual data points
  labs(
    title = "Unifrac Beta Dispersion by Group",
    x = "Group",
    y = "Distance to Centroid"
  ) +
  theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Do permanova with *Unifrac* on *labels* column:

```{r unifrac-labels_permanova}
#| label: unifrac-labels_permanova
# load the permanova with targets
targets::tar_load(nov_2023_unifrac_labels_permanova, store = store_path)
print(nov_2023_unifrac_labels_permanova)
```

Load the *pairwise* permanova on the same columns and filter only significant
values (P.adj < 0.05):

```{r unifrac-pairwise_labels_permanova}
#| label: unifrac-pairwise_labels_permanova
# load the permanova with targets
targets::tar_load(nov_2023_pairwise_unifrac_labels_permanova, store = store_path)
write.xlsx(nov_2023_pairwise_unifrac_labels_permanova, file = here(data_path, "nov_2023-bacteria_pairwise_unifrac_labels_permanova.xlsx"))

nov_2023_pairwise_unifrac_labels_permanova %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj) %>%
  dplyr::select(Comparison, P.adj, everything())
```

Download the full [nov_2023-bacteria_pairwise_unifrac_labels_permanova.xlsx](07-november_2023-bacteria/nov_2023-bacteria_pairwise_unifrac_labels_permanova.xlsx) table.

### Weighted UniFrac {#wunifrac}

#### PCoA plot by sample names

Now we will plot a Weighted Unifrac PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r wunifrac-pcoa}
#| label: wunifrac-pcoa
# retrieve the PCoA object from targets
targets::tar_load(nov_2023_wunifrac_pcoa_object, store = store_path)

pcoa_results <- nov_2023_wunifrac_pcoa_object$pcoa_results
percent_explained <- nov_2023_wunifrac_pcoa_object$percent_explained
pcoa_df <- nov_2023_wunifrac_pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r wunifrac-pcoa-plot}
#| label: wunifrac-pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Weighted Unifrac)") +
  ggplot2::labs(title = "PCoA of Weighted Unifrac Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by labels (sample replicates)

Visualize the same plot but grouping with labels (which should group sample replicates)

```{r wunifrac-pcoa-labels}
#| label: wunifrac-pcoa-labels
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Weigthed Unifrac)") +
  ggplot2::labs(title = "PCoA of Weighted Unifrac Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation from *reactor samples* from *default*.
*Box* samples are closer togheter but distant from *reactor* conditions and closer
to *default* than reactor.

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r wunifrac-scree-plot}
#| label: wunifrac-scree-plot
# Determine which cumulative eigenvalues to use
if ("Cum_corr_eig" %in% colnames(pcoa_results$values)) {
  cumulative_eig <- pcoa_results$values$Cum_corr_eig * 100
} else {
  cumulative_eig <- pcoa_results$values$Cumul_eig * 100
}

# Create a tibble for plotting
plot_data <- tibble(
  pe = cumulative_eig,
  axis = seq_along(cumulative_eig)
)

# Generate the plot
ggplot(plot_data, aes(x = axis, y = pe)) +
  geom_line() +
  geom_vline(xintercept = 2, color = "red") +
  labs(
    x = "Principal Coordinate Axis",
    y = "Cumulative Percentage of Variance Explained",
    title = "Scree Plot of PCoA (Weighted Unifrac)",
  ) +
  theme_minimal()
```

#### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

```{r wunifrac-nmds}
#| label: wunifrac-nmds
targets::tar_load(nov_2023_wunifrac_nmds_object, store = store_path)

# get the tibble object
nmds_tb <- nov_2023_wunifrac_nmds_object$nmds_tb
```
##### NMDS plot by sample names {#wunifrac-nmds-sample-names}

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

```{r wunifrac-nmds-plot}
#| label: wunifrac-nmds-plot
#| fig.width: 10
#| fig.height: 8

# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = sample_name, shape = condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Weigthed Unifrac)",
    color = NULL,
    shape = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

##### NMDS plot by labels (sample replicates) {#wunifrac-nmds-labels}

Now, let's visualize the NMDS plot but grouping with labels
(which should group sample replicates)

```{r wunifrac-nmds-labels}
#| label: wunifrac-nmds-labels

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = label)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Weigthed Unifrac)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```
