---
title: "November 2023 Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
    embed-resources: true  # Embed all resources directly into the HTML file
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(ampvis2)

store_path <- here::here("stores", "november_2023")
```

## Load metadata from november 2023

Let's load the metadata from the November 2023 experiments and try to do some analysis.

```{r load-metadata}
#| label: load-metadata
# load data from targets
targets::tar_load(nov_2023_metadata, store = store_path)

# check the new metadata
nov_2023_metadata
```

Load other phyloseq stuff

```{r load-phyloseq}
#| label: load-phyloseq
targets::tar_load(nov_2023_phyloseq_obj, store = store_path)
targets::tar_load(nov_2023_samples_data, store = store_path)
```

## Checking samples

### Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
# Create the bar plot
depth_plot <- ggplot2::ggplot(
  nov_2023_samples_data,
  aes(x = sampleID, y = sequencing_depth, fill = date_condition)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

### Permanova

Calculate the distance matrix using the *bray-curtis* method using a
*rarefaction* method as suggested by Patrick D. Schloss. Then
we will use the `adonis2` function from the `vegan` package to test if the technical
replicates are similar to each other. The formula should include the `sample_group`
as a factor to account for biological variability and `technical_rep` to assess
technical replicate similarity:

```{r tech_rep_permanova}
#| label: tech_rep_permanova
# load the permanova with targets
targets::tar_load(nov_2023_tech_rep_permanova, store = store_path)
print(nov_2023_tech_rep_permanova)
```

## Barplots

Let's create some barplots to visualize the relative abundance of taxa in the
dataset. Here are the available taxonomic ranks in the dataset:

```{r taxonomic-ranks}
#| label: taxonomic-ranks
#| description: List the available taxonomic ranks in the dataset.
phyloseq::rank_names(nov_2023_phyloseq_obj)
```

### By Phylum

Create a barplot of the relative abundance Phylum in the dataset. Sort samples
by sampleID.

```{r barplot-phylum}
#| label: barplot-phylum
#| description: Create a barplot of the relative abundance of Phylum in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum"
#| fig-align: center
# load the melted data
tar_load(nov_2023_melted_phylum, store = store_path)
taxrank <- "Phylum"

ggplot2::ggplot(nov_2023_melted_phylum, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Phylum (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance

```{r barplot-phylum-grouped}
#| label: barplot-phylum-grouped
#| description: Create a barplot of the relative abundance of Phylum in the dataset grouping taxa with less than 5% abundance.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum grouped (>5% abundance)"
taxrank <- "Phylum"
nov_2023_melted_phylum_grouped <- nov_2023_melted_phylum %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_phylum_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class

Create a barplot of the relative abundance Class in the dataset. Sort samples
by sampleID.

```{r barplot-class}
#| label: barplot-class
#| description: Create a barplot of the relative abundance of class in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class"
#| fig-align: center
# load the melted data
tar_load(nov_2023_melted_class, store = store_path)
taxrank <- "Class"

ggplot2::ggplot(nov_2023_melted_class, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance

```{r barplot-class-grouped}
#| label: barplot-class-grouped
#| description: Create a barplot of the relative abundance of Class in the dataset grouping taxa with less than 5% abundance.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class grouped (>5% abundance)"
taxrank <- "Class"
nov_2023_melted_class_grouped <- nov_2023_melted_class %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_class_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmaps

### By Phylum

Here are the heatmaps of the relative abundance of the top 10 most abundant
taxa at the Phylum level:

```{r heatmap-phylum}
#| label: heatmap-phylum
#| description: Create a heatmap of the relative abundance of the top 10 most abundant taxa at the Phylum level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 10 most abundant taxa at the Phylum level"
#| fig-align: center
tar_load(nov_23_heatmap_phylum, store = store_path)
print(nov_23_heatmap_phylum)
```

### By Class

Here are the heatmaps of the relative abundance of the top 20 most abundant
taxa at the Class level:

```{r heatmap-class}
#| label: heatmap-class
#| description: Create a heatmap of the relative abundance of the top 20 most abundant taxa at the Class level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 20 most abundant taxa at the Class level"
#| fig-align: center
tar_load(nov_23_heatmap_class, store = store_path)
print(nov_23_heatmap_class)
```

## Samples Rarefaction

Rarefaction is a technique used to estimate species richness or diversity
in a community by standardizing the number of samples or reads across different
samples. It involves subsampling the data to a common depth, allowing for a fair
comparison of diversity metrics between samples with varying sequencing depths.
The red line correspond to the lowest number of sequences present in samples: this
is the sequencing depth used in rarefaction step.

```{r rarefaction}
#| label: rarefaction
#| description: Perform rarefaction analysis on the OTU table
#|  and visualize the results.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Rarefaction curves with vegan"
# load calculated rarefaction data
tar_load(nov_23_rarecurve_df, store = store_path)
tar_load(rarefaction_depth, store = store_path)

# Add metadata to the rarecurve_df
nov_23_rarecurve_df <- nov_2023_metadata %>%
  select(sampleID, sample_name, date_condition) %>%
  inner_join(nov_23_rarecurve_df, by = c("sampleID" = "Site"))

# draw the plot
ggplot(nov_23_rarecurve_df, aes(x = Sample, y = Species, color = date_condition, group = sampleID)) +
  geom_line() +
  geom_vline(xintercept = rarefaction_depth, linetype = "dashed", color = "red") +  # Add vertical line
  labs(
    title = "Rarefaction Curves for All Samples",
    x = "Number of Individuals Sampled",
    y = "Number of Species Observed"
  ) +
  theme_bw()
```
