---
title: "November 2023 Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
    embed-resources: true  # Embed all resources directly into the HTML file
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(ampvis2)

store_path <- here::here("stores", "november_2023")
```

## Load metadata from november 2023

Let's load the metadata from the November 2023 experiments and try to do some analysis.

```{r load-metadata}
#| label: load-metadata
# load data from targets
targets::tar_load(nov_2023_metadata, store = store_path)

# check the new metadata
nov_2023_metadata
```

Load other phyloseq stuff

```{r load-phyloseq}
#| label: load-phyloseq
targets::tar_load(nov_2023_phyloseq_obj, store = store_path)
targets::tar_load(nov_2023_samples_data, store = store_path)
```

## Checking samples

### Display number of samples

Try to see sample depths and other information about the samples:

```{r sequencing-depth}
#| label: sequencing-depth
# Create the bar plot
depth_plot <- ggplot2::ggplot(
  nov_2023_samples_data,
  aes(x = sampleID, y = sequencing_depth, fill = date_condition)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Sequencing Depth per Sample",
       x = "Sample",
       y = "Number of Reads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Display the plot
print(depth_plot)
```

### Permanova

Calculate the distance matrix using the *bray-curtis* method using a
*rarefaction* method as suggested by Patrick D. Schloss. Then
we will use the `adonis2` function from the `vegan` package to test if the technical
replicates are similar to each other. The formula should include the `sample_group`
as a factor to account for biological variability and `technical_rep` to assess
technical replicate similarity:

```{r tech_rep_permanova}
#| label: tech_rep_permanova
# load the permanova with targets
targets::tar_load(nov_2023_tech_rep_permanova, store = store_path)
print(nov_2023_tech_rep_permanova)
```

## Barplots

Let's create some barplots to visualize the relative abundance of taxa in the
dataset. Here are the available taxonomic ranks in the dataset:

```{r taxonomic-ranks}
#| label: taxonomic-ranks
#| description: List the available taxonomic ranks in the dataset.
phyloseq::rank_names(nov_2023_phyloseq_obj)
```

### By Phylum

Create a barplot of the relative abundance Phylum in the dataset. Sort samples
by sampleID.

```{r barplot-phylum}
#| label: barplot-phylum
#| description: Create a barplot of the relative abundance of Phylum in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum"
#| fig-align: center
# load the melted data
tar_load(nov_2023_melted_phylum, store = store_path)
taxrank <- "Phylum"

ggplot2::ggplot(nov_2023_melted_phylum, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Phylum (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance

```{r barplot-phylum-grouped}
#| label: barplot-phylum-grouped
#| description: Create a barplot of the relative abundance of Phylum in the dataset grouping taxa with less than 5% abundance.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Phylum grouped (>5% abundance)"
taxrank <- "Phylum"
nov_2023_melted_phylum_grouped <- nov_2023_melted_phylum %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_phylum_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class

Create a barplot of the relative abundance Class in the dataset. Sort samples
by sampleID.

```{r barplot-class}
#| label: barplot-class
#| description: Create a barplot of the relative abundance of class in the dataset.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class"
#| fig-align: center
# load the melted data
tar_load(nov_2023_melted_class, store = store_path)
taxrank <- "Class"

ggplot2::ggplot(nov_2023_melted_class, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### By Class (>5%)

Same as before, but this time let's group the taxa with less than 5% abundance

```{r barplot-class-grouped}
#| label: barplot-class-grouped
#| description: Create a barplot of the relative abundance of Class in the dataset grouping taxa with less than 5% abundance.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Barplot of Class grouped (>5% abundance)"
taxrank <- "Class"
nov_2023_melted_class_grouped <- nov_2023_melted_class %>%
  dplyr::mutate(!!rlang::sym(taxrank) := ifelse(Abundance < 0.05, "Other Taxa", !!rlang::sym(taxrank))) %>%
  dplyr::group_by(sampleID, !!rlang::sym(taxrank)) %>%
  dplyr::summarise(Abundance = sum(Abundance), .groups = 'drop')

ggplot2::ggplot(nov_2023_melted_class_grouped, ggplot2::aes(x = sampleID, y = Abundance, fill = !!rlang::sym(taxrank))) +
  ggplot2::geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.5) +
  # need a sorted object like the one used in agglomerate_by_taxa
  ggplot2::scale_x_discrete(labels = nov_2023_metadata$date_condition) +
  ggplot2::labs(
    title = glue::glue("Microbial composition at {taxrank} level grouped (>5% abundance)"),
    x = "Sample",
    y = "Relative Abundance"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmaps

### By Phylum

Here are the heatmaps of the relative abundance of the top 10 most abundant
taxa at the Phylum level:

```{r heatmap-phylum}
#| label: heatmap-phylum
#| description: Create a heatmap of the relative abundance of the top 10 most abundant taxa at the Phylum level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 10 most abundant taxa at the Phylum level"
#| fig-align: center
tar_load(nov_2023_heatmap_phylum, store = store_path)
print(nov_2023_heatmap_phylum)
```

### By Class

Here are the heatmaps of the relative abundance of the top 20 most abundant
taxa at the Class level:

```{r heatmap-class}
#| label: heatmap-class
#| description: Create a heatmap of the relative abundance of the top 20 most abundant taxa at the Class level.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Heatmap of the top 20 most abundant taxa at the Class level"
#| fig-align: center
tar_load(nov_2023_heatmap_class, store = store_path)
print(nov_2023_heatmap_class)
```

## Samples Rarefaction

Rarefaction is a technique used to estimate species richness or diversity
in a community by standardizing the number of samples or reads across different
samples. It involves subsampling the data to a common depth, allowing for a fair
comparison of diversity metrics between samples with varying sequencing depths.
The red line correspond to the lowest number of sequences present in samples: this
is the sequencing depth used in rarefaction step.

```{r rarefaction}
#| label: rarefaction
#| description: Perform rarefaction analysis on the OTU table
#|  and visualize the results.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Rarefaction curves with vegan"
# load calculated rarefaction data
tar_load(nov_2023_rarecurve_df, store = store_path)
tar_load(nov_2023_rarefaction_depth, store = store_path)

# Add metadata to the rarecurve_df
nov_2023_rarecurve_df <- nov_2023_metadata %>%
  select(sampleID, sample_name, date_condition) %>%
  inner_join(nov_2023_rarecurve_df, by = c("sampleID" = "Site"))

# draw the plot
ggplot(nov_2023_rarecurve_df, aes(x = Sample, y = Species, color = date_condition, group = sampleID)) +
  geom_line() +
  geom_vline(xintercept = nov_2023_rarefaction_depth, linetype = "dashed", color = "red") +  # Add vertical line
  labs(
    title = "Rarefaction Curves for All Samples",
    x = "Number of Individuals Sampled",
    y = "Number of Species Observed"
  ) +
  theme_bw()
```

## Alpha diversity

Alpha diversity measures the diversity within a single sample or community.
It provides insights into the richness and evenness of species in a sample,
helping to compare diversity between samples or track changes over time.

### By sample_name

```{r alpha-diversity_sample_name}
#| label: alpha-diversity_sample_name
#| description: Calculate and visualize alpha diversity metrics for the samples.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Alpha diversity metrics for the samples"
tar_load(nov_2023_alpha_diversity_by_sample_name, store = store_path)
print(nov_2023_alpha_diversity_by_sample_name)
```

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on *shannon* metric:

```{r kruscal_shannon_sample_name}
#| label: kruscal_shannon_sample_name
#| description: Perform Kruskal-Wallis tests on the Shannon ~ sample_name
tar_load(nov_2023_kruscal_shannon_sample_name, store = store_path)
print(nov_2023_kruscal_shannon_sample_name)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test to find significant
differences between pairs:

```{r dunn_shannon_sample_name}
#| label: dunn_shannon_sample_name
tar_load(nov_2023_dunn_shannon_sample_name, store = store_path)
nov_2023_dunn_shannon_sample_name$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

### By date_condition

```{r alpha-diversity_date_condition}
#| label: alpha-diversity_date_condition
#| description: Calculate and visualize alpha diversity metrics for date-condition.
#| fig-width: 10
#| fig-height: 8
#| fig-caption: "Alpha diversity metrics for date-condition"
tar_load(nov_2023_alpha_diversity_by_date_condition, store = store_path)
print(nov_2023_alpha_diversity_by_date_condition)
```

Perform Kruskal-Wallis tests to determine if the alpha diversity metrics are
significantly different between the groups. Run on shannon metric:

```{r kruscal_shannon_date_condition}
#| label: kruscal_shannon_date_condition
#| description: Perform Kruskal-Wallis tests on the Shannon ~ date_condition
tar_load(nov_2023_kruscal_shannon_date_condition, store = store_path)
print(nov_2023_kruscal_shannon_date_condition)
```

This means that *at least* one group is significantly different in terms of
Shannon diversity. Let's do a post-hoc test with Dunn's test to find significant
differences between pairs:

```{r dunn_shannon_date_condition}
#| label: dunn_shannon_date_condition
tar_load(nov_2023_dunn_shannon_date_condition, store = store_path)
nov_2023_dunn_shannon_date_condition$res %>%
  dplyr::filter(P.adj < 0.05) %>%
  dplyr::arrange(P.adj)
```

## Beta diversity

### PCoA plot

**Principal Coordinates Analysis (PCoA)** is an ordination technique used to visualize patterns in complex, high-dimensional data by projecting samples into a lower-dimensional space based on a distance matrix.

#### PCoA plot by sample names

Now we will plot a PCoA plot to visualize the technical replicates. See
[CC186](https://www.youtube.com/watch?v=G5Qckqq5Erw) for more information.

```{r pcoa}
#| label: pcoa
# retrieve the PCoA object from targets
targets::tar_load(nov_2023_pcoa_object, store = store_path)

pcoa_results <- nov_2023_pcoa_object$pcoa_results
percent_explained <- nov_2023_pcoa_object$percent_explained
pcoa_df <- nov_2023_pcoa_object$pcoa_df
```

In the following pictures we are grouping by technical replicates

```{r pcoa-plot}
#| label: pcoa-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot_rep <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = sample_name)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(pcoa_plot_rep)
```

#### PCoA plot by date and condition (sample replicates)

Visualize the same plot but grouping with date and condition (which should group sample replicates)

```{r pcoa-date-condition}
#| label: pcoa-date-condition
# try to annotate plot
pretty_pe <- format(round(percent_explained, digits = 1), nsmall = 1, trim = TRUE)
labels <- c(glue::glue("PCo Axis 1 ({pretty_pe[1]}%)"),
            glue::glue("PCo Axis 2 ({pretty_pe[2]}%)"))

# plot the PCoA
pcoa_plot <- ggplot2::ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = date_condition)) +
  ggplot2::geom_point(size = 4, alpha = 0.5) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCoA Plot on Distance Metrics (Bray-Curtis)") +
  ggplot2::labs(title = "PCoA of Bray-Curtis Distance Matrix",
       color = NULL,
       x = labels[1],
       y = labels[2]) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

pcoa_plot
```

In this image there's a clear separation from *reactor samples* from *default*.
*Box* samples are closer togheterbut distant from *reactor* conditions.

#### Scree plot

The scree plot is a graphical representation of the eigenvalues of the principal
components in a PCA or PCoA analysis. It helps visualize the proportion of variance
explained by each principal component, allowing you to determine the optimal
number of components to retain. The red line in the plot is at y=2, which is the
number of components used when plotting a 2D PCoA plot.

```{r scree-plot}
#| label: scree-plot
# Determine which cumulative eigenvalues to use
if ("Cum_corr_eig" %in% colnames(pcoa_results$values)) {
  cumulative_eig <- pcoa_results$values$Cum_corr_eig * 100
} else {
  cumulative_eig <- pcoa_results$values$Cumul_eig * 100
}

# Create a tibble for plotting
plot_data <- tibble(
  pe = cumulative_eig,
  axis = seq_along(cumulative_eig)
)

# Generate the plot
ggplot(plot_data, aes(x = axis, y = pe)) +
  geom_line() +
  geom_vline(xintercept = 2, color = "red") +
  labs(
    x = "Principal Coordinate Axis",
    y = "Cumulative Percentage of Variance Explained",
    title = "Scree Plot of PCoA"
  ) +
  theme_minimal()
```

### NMDS plot

Non-metric Multidimensional Scaling (**NMDS**) is an ordination technique used to visualize the similarity or dissimilarity between samples in complex, multidimensional datasets. Unlike metric methods, NMDS focuses on preserving the rank order of distances rather than the actual distance values, making it particularly useful for ecological and environmental data where relationships may be non-linear.

```{r nmds}
#| label: nmds
targets::tar_load(nov_2023_nmds_object, store = store_path)

# get the tibble object
nmds_tb <- nov_2023_nmds_object$nmds_tb
jitter <- 0.05
```
#### NMDS plot by sample names

Let's plot the NMDS plot focusing on the technical replicates.
see [CC187](https://www.youtube.com/watch?v=h7OrVmT7Ja8&ab_channel=RiffomonasProject)
for more information.

```{r nmds-plot}
#| label: nmds-plot
# count the number of levels in the group column
num_levels <- length(unique(nov_2023_metadata$sample_group))

# Get the color palette
color_palette <- Polychrome::palette36.colors(num_levels)
names(color_palette) <- NULL  # Remove any names from the color palette

# Use palette36 from Polychrome, which supports up to 36 colors
nmds_plot_rep <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = sample_name, shape = condition)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL,
    shape = NULL) +
  ggplot2::scale_color_manual(values = color_palette) +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot_rep)
```

::: {.callout-warning}
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

#### NMDS plot by date and condition (sample replicates)

Now, let's visualize the NMDS plot but grouping with date and condition
(which should group sample replicates)

```{r nmds-date-condition}
#| label: nmds-date-condition

# Plot the NMDS
nmds_plot <- ggplot2::ggplot(
    nmds_tb,
    ggplot2::aes(x = NMDS1, y = NMDS2, color = date_condition)) +
  ggplot2::geom_jitter(size = 4, alpha = 0.5, width = jitter, height = jitter) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "NMDS Plot on distance metrics (Bray-Curtis)",
    color = NULL) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::theme(
    legend.position = "bottom",       # Position the legend at the bottom
    legend.box = "horizontal",        # Arrange legend items horizontally without a title
  )

print(nmds_plot)
```

::: {.callout-warning}
I've applied `0.05` of jitter (random noise) to points in order
to clearly see them in the image.
:::

### Between group distances

#### sample_name distances

Plot distances by sample names:

```{r sample_name_distances}
#| label: sample_name_distances
#| title: Distance matrix by sample names
#| description: Plot the distance matrix by sample names
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_bray_distance_by_sample_name_plot, store = store_path)
print(nov_2023_bray_distance_by_sample_name_plot)
```

#### date_condition distances

Plot distances by date_condition:

```{r date_condition_distances}
#| label: date_condition_distances
#| title: Distance matrix by date_condition
#| description: Plot the distance matrix by date_condition
#| fig.width: 10
#| fig.height: 8
#| fig.align: center
tar_load(nov_2023_bray_distance_by_date_condition_plot, store = store_path)
print(nov_2023_bray_distance_by_date_condition_plot)
```
