---
title: "Bacteria - November 2023 Sampling Completeness"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
cache: false
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 4
    embed-resources: true  # Embed all resources directly into the HTML file
knitr:
  opts_chunk:
    dpi: 300
editor: source
---

```{r setup}
#| label: setup
#| output: false
library(targets)
library(here)
library(tidyverse)
library(ggplot2)
library(gt)
library(glue)
library(openxlsx)

# where I can find artifacts
store_path <- here::here("stores", "november_2023-sampling_completeness")

# a folder in which store data files
data_path <- here::here("analysis", "09-november_2023-sampling_completeness")

if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  cat("Directory created:", data_path, "\n")
}

theme_set(theme_bw(base_size = 14))
```

## Sampling completeness

The rarefaction curves shown in the `november_2023-bacteria` report suggest that
sequencing depth was sufficient to capture most of the bacterial diversity in each
sample, but a visual inspection of curves flattening out is not, by itself, a
quantitative argument. Here we use `iNEXT` (Hill number `q = 0`, i.e. species
richness) to compare, for each sample:

- the observed/extrapolated richness at the rarefaction depth actually used for the
  diversity analyses (`nov_2023_rarefaction_depth`, the minimum sequencing depth across
  the November 2023 samples), and
- the asymptotic richness estimated by the Chao1 estimator, i.e. the richness expected
  at infinite sequencing effort.

The ratio of the two — the *coverage ratio* — quantifies how close each sample already
is to its estimated total diversity. A coverage ratio below 90% flags samples for which
a substantial share of the estimated diversity may not have been captured at the chosen
rarefaction depth.

```{r load-completeness}
#| label: load-completeness
targets::tar_load(nov_2023_sampling_completeness, store = store_path)
targets::tar_load(nov_2023_completeness_group_summary, store = store_path)
targets::tar_load(nov_2023_rarefaction_depth, store = store_path)

write.xlsx(
  nov_2023_sampling_completeness,
  file = here(data_path, "nov_2023-sampling_completeness.xlsx")
)
```

### Per-sample coverage table

```{r completeness-table}
#| label: completeness-table
nov_2023_sampling_completeness %>%
  gt::gt() %>%
  gt::fmt_number(
    columns = c(Observed_or_estimated_richness, Asymptotic_Chao1_estimate),
    decimals = 1
  ) %>%
  gt::fmt_percent(columns = Coverage_ratio, decimals = 1) %>%
  gt::data_color(
    columns = Flag_below_90pct,
    fn = function(x) ifelse(x, "#f4a3a3", "#a3d9a5")
  ) %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(rows = Flag_below_90pct)
  ) %>%
  gt::tab_header(
    title = "Sampling completeness per sample",
    subtitle = glue::glue(
      "Coverage ratio at a reference depth of {format(nov_2023_rarefaction_depth, big.mark = ',')} reads"
    )
  )
```

Download the full [nov_2023-sampling_completeness.xlsx](09-november_2023-sampling_completeness/nov_2023-sampling_completeness.xlsx) table.

### Summary

```{r completeness-summary}
#| label: completeness-summary
#| output: asis
n_samples <- nrow(nov_2023_sampling_completeness)
n_above_90 <- sum(!nov_2023_sampling_completeness$Flag_below_90pct)
pct_above_90 <- round(100 * n_above_90 / n_samples, 1)

low_coverage_groups <- nov_2023_completeness_group_summary %>%
  dplyr::filter(n_below_90 > 0) %>%
  dplyr::arrange(mean_coverage)

low_coverage_text <- if (nrow(low_coverage_groups) > 0) {
  groups_list <- glue::glue_collapse(
    glue::glue(
      "{low_coverage_groups$Group} (mean coverage {round(100 * low_coverage_groups$mean_coverage, 1)}%, ",
      "{low_coverage_groups$n_below_90}/{low_coverage_groups$n_samples} sample(s) below 90%)"
    ),
    sep = "; "
  )
  glue::glue("The following group(s) showed lower coverage: {groups_list}.")
} else {
  "No experimental group showed samples below the 90% coverage threshold."
}

cat(glue::glue(
  "At the reference sequencing depth of {format(nov_2023_rarefaction_depth, big.mark = ',')} reads, ",
  "{n_above_90} out of {n_samples} samples ({pct_above_90}%) reached at least 90% of their ",
  "Chao1-estimated asymptotic richness, supporting the conclusion that sequencing depth was ",
  "sufficient to characterize the majority of the detectable bacterial diversity. {low_coverage_text}"
))
```

### Coverage plot

```{r completeness-plot}
#| label: completeness-plot
#| fig-width: 8
#| fig-height: 6
ggplot(
  nov_2023_sampling_completeness,
  aes(x = reorder(Assemblage, Coverage_ratio), y = Coverage_ratio, fill = Flag_below_90pct)
) +
  geom_col(color = "black") +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "red") +
  coord_flip() +
  scale_fill_manual(
    values = c(`FALSE` = "steelblue", `TRUE` = "firebrick"),
    labels = c(`FALSE` = "≥ 90%", `TRUE` = "< 90%"),
    name = "Coverage"
  ) +
  labs(
    title = "Sampling completeness per sample",
    x = "Sample",
    y = "Coverage ratio (richness at reference depth / Chao1 estimate)"
  ) +
  theme_bw(base_size = 14)
```
